---
title: "Hands-on Exercise 10A"
author: "Kock Si Min"
date: October 31, 2024
date-modified: "last-modified"
toc: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

# Processing and Visualising Flow Data

## 10.1 Overview

Spatial interaction represents the flow of people, material or information between locations in a geographical space. It encompasses everything from freight shipments, energy flows, global trade in rare antiquities, flight schedules, rush hour woes and pedestrian foot traffic.

Each spatial interaction, as an analogy for a set of movements, is composed of a discrete origin/destination pair. Each pair can be presented as a cell in a matrix where rows are related to the locations (centroids) of origin while columns are related to locations (centroids) of destination. Such a matrix is known as an origin/destination (OD) matrix or spatial interaction matrix.

In this hands-on exercise, I will build an OD matrix by using the [*Passenger Volume by Origin Destination Bus Stops*](https://datamall.lta.gov.sg/content/datamall/en/search_datasets.html?searchText=passenger%20origin) data set downloaded from [LTA DataMall](https://datamall.lta.gov.sg/content/datamall/en.html).

By the end of the hands-on exercise, I would have:

-   imported and extracted OD data for a selected time interval,

-   imported and saved geospatial data (i.e. bus stops and mpsz) into sf tibble data frame objects,

-   populated planning subzone code into bus stops sf tibble data frame,

-   constructed desired lines in the geospatial data from the OD data, and

-   visualised passenger volume by origin and destination bus stops by using the desired lines data.

## 10.2 Getting Started

Before we start, we will need to install and load the necessary R packages into the R environment:

-   [sf](https://r-spatial.github.io/sf/) for importing, integrating, processing and transforming geospatial data.

-   [tidyverse](https://www.tidyverse.org/) for importing, integrating, wrangling and visualising data.

-   [tmap](https://r-tmap.github.io/tmap/) for creating elegent and cartographic quality thematic maps.

-   [stplanr](https://docs.ropensci.org/stplanr/) provides functions for solving common problems in transport planning and modelling such as downloading and cleaning transport datasets; creating geographic “desire lines” from origin-destination (OD) data; route assignment, locally and interfaces to routing services such as CycleStreets.net; calculation of route segment attributes such as bearing and aggregate flow; and ‘travel watershed’ analysis.

-   [DT](https://rstudio.github.io/DT/) provides an R interface to the JavaScript library DataTables. R data objects (matrices or data frames) can be displayed as tables on HTML pages, and DataTables provides filtering, pagination, sorting, and many other features in the tables.

```{r}
pacman::p_load(sf,tidyverse,tmap,stplanr, DT)
```

## 10.3 Preparing the Flow Data

### 10.3.1 Importing the OD data

The code chunk below uses *read_csv()* of **readr** package to import the *Passenger Volume by Origin Destination Bus Stops* dataset as a tibble data frame into the R environment:

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202210.csv")
```

We display the *odbus* tibble data table using the code chunk below:

```{r}
glimpse(odbus)
```

We note that the values in ORIGIN_PT_CODE and DESTINATION_PT_CODE are in numeric data type. We use the code chunk below to convert them into character data type:

```{r}
odbus$ORIGIN_PT_CODE <- as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <- as.factor(odbus$DESTINATION_PT_CODE)
```

### 10.3.2 Extracting the study area

For the purpose of this exercise, we will extract commuter flows on weekdays and between 6 am to 9am:

```{r}
odbus6_9 <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 6 &
           TIME_PER_HOUR <=9) %>%
  group_by(ORIGIN_PT_CODE,
           DESTINATION_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS))
```

We look at the contents of *odbus6_9* using the code chunk below:

```{r}
datatable(odbus6_9)
```

We will save the output as a new rds file and load it into the R environment for further use:

```{r}
#| eval: false
write_rds(odbus6_9,"data/rds/odbus6_9.rds")
```

```{r}
odbus6_9 <- read_rds("data/rds/odbus6_9.rds")
```

## 10.4 Working with Geospatial Data

For the purpose of this exercise, two geospatial data will be used:

-   BusStop: Location of bus stops as at last quarter of 2022

-   MPSZ-2019: sub-zone boundary of URA Master Plan 2019

Both files are in ESRI shapefile format.

### 10.4.1 Importing geospatial data

Note that when loading the geospatial files below, we apply st_transform() to ensure that the data files have the right EPSG code 3414.

```{r}
busstop <- st_read(dsn = "data/geospatial",
                  layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

We note that BusStop has a point form geometry while mpsz has a multipolygon geometry.

We will save the BusStop and mpsz as new rds files and load them into the R environment for further use:

```{r}
#| eval: false
write_rds(busstop,"data/rds/busstop.rds")
write_rds(mpsz,"data/rds/mpsz.rds")
```

```{r}
busstop <- read_rds("data/rds/busstop.rds")
mpsz <- read_rds("data/rds/mpsz.rds")
```

## 10.5 Geospatial data wrangling

### 10.5.1 Combining busstop and mpsz

The code chunk below populates the planning subzone code of mpsz sf data frame into the busstop sf data frame:

```{r}
busstop_mpsz <- st_intersection(busstop,mpsz) %>%
  select(BUS_STOP_N,SUBZONE_C) %>%
  st_drop_geometry()
```

::: callout-note
-   `st_intersection()` is used to perform point and polygon overlay and the output will be in point sf object.

-   `select()` of dplyr package is then use to retain only BUS_STOP_N and SUBZONE_C in the busstop_mpsz sf data frame.

-   five bus stops are excluded in the resultant data frame because they are outside of Singapore's boundary.
:::

```{r}
datatable(busstop_mpsz)
```

We save the output as a new rds file and load it into the R environment for further use:

```{r}
#| eval: false
write_rds(busstop_mpsz,"data/rds/busstop_mpsz.rds")
```

```{r}
busstop_mpsz <- read_rds("data/rds/busstop_mpsz.rds")
```

Next, we will append the planning subzone code from the busstop_mpsz data frame onto the odbus6_9 data frame:

```{r}
od_data <- left_join(odbus6_9,busstop_mpsz,
                     by= c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
  rename(ORIGIN_BS = ORIGIN_PT_CODE,
         ORIGIN_SZ = SUBZONE_C,
         DESTIN_BS = DESTINATION_PT_CODE)
```

Before continuing, it is a good practice to check for duplicate records:

```{r}
duplicate <- od_data %>%
  group_by_all() %>%
  filter(n() > 1) %>%
  ungroup()
```

```{r}
glimpse(duplicate)
```

We note that there are 976 duplicated records, the code chunk below is used to retain just the unique records:

```{r}
od_data <- unique(od_data)
```

We confirm if the duplicate record issue has been fully addressed by reusing the code chunk below:

```{r}
duplicate <- od_data %>%
  group_by_all() %>%
  filter(n() > 1) %>%
  ungroup()
duplicate
```

We note that there are now 0 duplicate records.

Next, we will update the od_data data frame with the planning subzone codes:

```{r}
od_data <- left_join(od_data, busstop_mpsz,
                     by = c("DESTIN_BS" = "BUS_STOP_N"))
```

We check for and remove duplicate data as above:

```{r}
duplicate <- od_data %>%
  group_by_all() %>%
  filter(n() > 1) %>%
  ungroup()

duplicate
```

There are 1256 duplicate records.

```{r}
od_data <- unique(od_data)
```

We double check for duplicate records:

```{r}
duplicate <- od_data %>%
  group_by_all() %>%
  filter(n() > 1) %>%
  ungroup()
duplicate
```

We note that there are now 0 duplicate records.

```{r}
od_data <- od_data %>%
  rename(DESTIN_SZ = SUBZONE_C) %>%
  drop_na() %>%
  group_by(ORIGIN_SZ,DESTIN_SZ) %>%
  summarise(MORNING_PEAK = sum(TRIPS))
```

We save the output as a new rds file and load into the R environment for further analysis:

```{r}
#| eval: false
write_rds(od_data,"data/rds/od_data_fii.rds")
```

```{r}
od_data_fii <- read_rds("data/rds/od_data_fii.rds")
```

## 10.6 Visualising Spatial Interaction

In this section, we will prepare **desire lines - straight lines between the origin and destination and represent where people would go if they were not constrained by the route network** - using [**stplanr**](#0)package.

### 10.6.1 Removing intra-zonal flows

We will not plot the intra-zonal flows and remove them using the code chunk below:

```{r}
od_data_fij <- od_data[od_data$ORIGIN_SZ != od_data$DESTIN_SZ,]
```

We save this output as a new rds file and load it into the R environment for further analysis:

```{r}
#| eval: false
write_rds(od_data_fij,"data/rds/od_data_fij.rds")
```

```{r}
od_data_fij <- read_rds("data/rds/od_data_fij.rds")
```

### 10.6.2 Creating desire lines

In this code chunk below, we will use `od2line()` of **stplanr** package is used to create the desire lines.

```{r}
flowLine <- od2line(flow = od_data_fij,
                    zones = mpsz,
                    zone_code = "SUBZONE_C")
```

We save the output as a new rds file and load it into the R environment:

```{r}
#| eval: false
write_rds(flowLine, "data/rds/flowLine.rds")
```

```{r}
flowLine <- read_rds("data/rds/flowLine.rds")
```

### 10.6.3 Visualising the desire lines

```{r}
tm_shape(mpsz) +
  tm_polygons()+
  flowLine %>%
  tm_shape() +
  tm_lines(lwd = "MORNING_PEAK",
           style = "quantile",
           scale = c(0.1,1,3,5,7,10),
           n = 6,
           alpha = 0.3)
```

::: callout-warning
Note that the rendering process takes more time during to the transparency argument i.e. alpha
:::

When the flow data is messy and highly skewed like the one above, it is wiser to focus on selected flows i.e. flows that are equal or greater than 5000 as shown below:

```{r}
tm_shape(mpsz)+
  tm_polygons()+
  flowLine %>%
  filter(MORNING_PEAK >= 5000) %>%
  tm_shape()+
  tm_lines(lwd = "MORNING_PEAK",
           style = "quantile",
           scale = c(0.1,1,3,5,7,10),
           n = 6,
           alpha = 0.3)
```
