---
title: "Hands-on Exercise 9"
author: "Kock Si Min"
date: October 26, 2024
date-modified: "last-modified"
toc: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

# Processing and Visualising Flow Data

## 10.1 Overview

Spatial interaction represents the flow of people, material or information between locations in a geographical space. It encompasses everything from freight shipments, energy flows, global trade in rare antiquities, flight schedules, rush hour woes and pedestrian foot traffic.

Each spatial interaction, as an analogy for a set of movements, is composed of a discrete origin/destination pair. Each pair can be presented as a cell in a matrix where rows are related to the locations (centroids) of origin while columns are related to locations (centroids) of destination. Such a matrix is known as an origin/destination (OD) matrix or spatial interaction matrix.

In this hands-on exercise, I will build an OD matrix by using the [*Passenger Volume by Origin Destination Bus Stops*](https://datamall.lta.gov.sg/content/datamall/en/search_datasets.html?searchText=passenger%20origin) data set downloaded from [LTA DataMall](https://datamall.lta.gov.sg/content/datamall/en.html).

By the end of the hands-on exercise, I would have:

-   imported and extracted OD data for a selected time interval,

-   imported and saved geospatial data (i.e. bus stops and mpsz) into sf tibble data frame objects,

-   populated planning subzone code into bus stops sf tibble data frame,

-   constructed desired lines in the geospatial data from the OD data, and

-   visualised passenger volume by origin and destination bus stops by using the desired lines data.

## 10.2 Getting Started

Before we start, we will need to install and load the necessary R packages into the R environment:

-   [sf](https://r-spatial.github.io/sf/) for importing, integrating, processing and transforming geospatial data.

-   [tidyverse](https://www.tidyverse.org/) for importing, integrating, wrangling and visualising data.

-   [tmap](https://r-tmap.github.io/tmap/) for creating elegent and cartographic quality thematic maps.

-   [stplanr](https://docs.ropensci.org/stplanr/) provides functions for solving common problems in transport planning and modelling such as downloading and cleaning transport datasets; creating geographic “desire lines” from origin-destination (OD) data; route assignment, locally and interfaces to routing services such as CycleStreets.net; calculation of route segment attributes such as bearing and aggregate flow; and ‘travel watershed’ analysis.

-   [DT](https://rstudio.github.io/DT/) provides an R interface to the JavaScript library DataTables. R data objects (matrices or data frames) can be displayed as tables on HTML pages, and DataTables provides filtering, pagination, sorting, and many other features in the tables.

```{r}
pacman::p_load(sf,tidyverse,tmap,stplanr, DT, XML, xml2)
```

## 10.3 Preparing the Flow Data

### 10.3.1 Importing the OD data

The code chunk below uses *read_csv()* of **readr** package to import the *Passenger Volume by Origin Destination Bus Stops* dataset as a tibble data frame into the R environment:

```{r}
#| eval: false
odbus <- xmlParse("data/aspatial/PassengerVolumeByOriginDestinationBusStops.xml")
```

::: callout-note

**skip**

:   Number of lines to skip before reading data. If `comment` is supplied any commented lines are ignored *after* skipping.
:::
