---
title: "Take-home Exercise 3"
author: "Kock Si Min"
date: October 26, 2024
date-modified: "last-modified"
toc: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

# **Take-home Exercise 3: Predicting HDB Resale Prices with Geographically Weighted Machine Learning Methods**

## 1. Setting the Scene

Housing is a sign of household wealth. It is a major investment for most people and its price is affected by several factors that can be global in nature such as the general economy of a country, inflation rate or are factors specific to the nature of the property which can be further divided to structural and locational factors. Structural variables are related to the property such as the size, fitting, tenure while the locational variables are related to the neighbourhood of the property such as proximity to childcare services, public transport and shopping centres.

Conventionally, predictive models for housing resale prices were built by using the Ordinary Least Square (OLS) method. However, this method failed to take into consideration that spatial autocorrelation and spatial heterogeneity exist in geographic data sets such as housing transactions. With the existence of spatial autocorrelation, the OLS estimation of predictive housing resale pricing models could lead to biased, inconsistent, or inefficient results (Anselin 1998). In view of this limitation, **Geographical Weighted Models** were introduced to better calibrate predictive models for housing resale prices.

## 2. The Task

In this take-home exercise, we will calibrate a predictive model to predict HDB resale prices between July-September 2024 by using HDB resale transaction records in 2023.

## 3. The Data

The following datasets will be used:

-   **Aspatial Data:**

    -   [HDB Resale Flat Prices](https://data.gov.sg/datasets/d_8b84c4ee58e3cfc0ece0d773c8ca6abc/view) from [Data.gov.sg](https://data.gov.sg/): A list of HDB resale transacted prices in Singapore from Jan 2017 onwards. In this exercise, our focus will be on 3-room flat types, given the [increasing resale price of 3-room HDB flats in recent years.](https://www.asiaone.com/money/more-3-room-resale-flats-change-hands-800k-and-higher-2024)

-   **Geospatial dataset**:

    -   URA's Master Plan 2019 Subzone Boundary from [Data.gov.sg](https://data.gov.sg/)

    -   Locational factors with geographic coordinates from [Data.gov.sg](https://data.gov.sg/)

        | S/N | Data                                                           |
        |-----|----------------------------------------------------------------|
        | 1   | **Eldercare**: List of eldercare in Singapore                  |
        | 2   | **Hawker Centre**: List of hawker centres in Singapore         |
        | 3   | **Parks:** List of parks in Singapore                          |
        | 4   | **Supermarket**: List of supermarkets in Singapore             |
        | 5   | **CHAS clinics**: List of CHAS clinics in Singapore            |
        | 6   | **Childcare service:** List of childcare services in Singapore |
        | 7   | **Kindergartens**: List of kindergartens in Singapore          |

    -   Locational factors with geographic coordinates from [Datamall.lta.gov.sg](https://datamall.lta.gov.sg/content/datamall/en.html):

        | S/N | Data                                                                                      |
        |-------------------|-----------------------------------------------------|
        | 1   | **MRT** data is a list of MRT/LRT stations in Singapore with the station names and codes. |
        | 2   | **Bus stops** data is a list of bus stops in Singapore                                    |

    -   Locational factors without geographic coordinates from [Data.gov.sg](https://data.gov.sg/) and other sources:

        | S/N | Data                                                                                                                   | Source                                                                                        |
        |------------------|------------------------------|------------------------|
        | 1   | **Primary school** data is extracted from the list on General information of schools                                   | [Data.gov.sg](https://data.gov.sg/)                                                           |
        | 2   | **Good primary schools** is a list of primary schools that are ordered in ranking in terms of popularity for year 2024 | [Local Salary Forum](https://www.salary.sg/2024/best-primary-schools-2024-by-popularity/)     |
        | 3   | **CBD** coordinates obtained from Google                                                                               | Google                                                                                        |
        | 4   | **Shopping malls** data is a list of Shopping malls in Singapore                                                       | obtained from [Wikipedia](https://en.wikipedia.org/wiki/List_of_shopping_malls_in_Singapore). |

## 4. Installing and Loading R packages

We will install and load the following R packages:

-   [httr](https://httr.r-lib.org/): allow R to talk to http

-   [rvest](https://rvest.tidyverse.org/): used to crawl websites, less flexible than the version in python but easy to start with

-   [jsonlite](https://cran.r-project.org/web/packages/jsonlite/): to communicate with one map API and convert data into dataframe format

```{r}
pacman::p_load(tidyverse, sf, GWmodel, SpatialML, tmap, rsample, Metrics, httr, jsonlite, rvest, fastDummies, corrplot)
```

## 5. Aspatial Data Wrangling

### 5.1 HDB Resale Prices Data Wrangling

We will load the HDB Resale Flat Prices data using the code chunk below and we filter the data to (i) retain data from 2023 onwards as the intent of the take-home exercise is to predict HDB resale prices in July to September 2024 using 2023 data and (ii) focus on just 3-room flat types.A glimpse at the data indicates that there are 11642 observations across 11 variables:

```{r}
resale <- read_csv("data/aspatial/hdbresale.csv") %>%
  filter(month >= "2023-01" & month <= "2024-09" & flat_type == "3 ROOM")
```

A glimpse at the data indicates that there are 11642 observations across 11 variables:

```{r}
glimpse(resale)
```

### 5.1.1 Creating new variables

Based on the output above, we also note that the address of the HDB flats have been split into "block", "street_name" and has no postal code. To facilitate the development of geographically weighted predictive models, we will need to combine the "block" and "street_name" to form a new field "address" and utilise [One Map API](https://www.onemap.gov.sg/apidocs/) to carry out reverse geocoding i.e. to return x, y coordinates for the address that is passed through the system.

In the code chunk below, we will create "address" as a new field and as we do so, we also split the "remaining_lease" field which is current in the XX years and YY months format to "remaining_lease_yr" and "remaining_lease_mth" respectively before converting the "remaining_lease_yr" to months by multiplying it by 12 and adding on to the "remaining_lease_mth" field to obtain the remaining lease in terms of months:

```{r}
resale <- resale %>%
  mutate(address = paste(block, street_name)) %>%
  mutate(remaining_lease_yr = as.integer(str_extract(remaining_lease, "^[0-9]+"))) %>%
  mutate(remaining_lease_mth = as.integer(str_extract(remaining_lease, "(?<=years, )[0-9]+"))) %>%
  mutate(remaining_lease_mth = ifelse(is.na(remaining_lease_mth), 0, remaining_lease_mth)) %>%
  mutate(remaining_lease_mth = remaining_lease_yr * 12 + remaining_lease_mth) %>%
  mutate(age = 2024 - lease_commence_date)
```

We double check that the above code was run properly using the code chunk below:

```{r}
na_rows <- resale[rowSums(is.na(resale)) > 0, ]
na_rows
```

We note that there are no NA values.

### 5.1.2 Reverse geocoding using One Map API

To carry reverse geocoding, we will need to create a list of unique addresses from the *resale* dataset. However, it is important to note that we should not push too much data i.e. 10,000 to the system as it will result in a longer lag time before results can be returned. We hence first split the *resale* data so that we divide the 11642 observations into two:

```{r}
resale1 <- resale %>%
  filter(month <= "2023-10")

resale2 <- resale %>%
  filter(month > "2023-10" & month <= "2024-09")
```

We utilise the *resale1* and *resale2* datasets to generate two lists of unique addresses:

```{r}
add_list1 <- sort(unique(resale1$address))
add_list2 <- sort(unique(resale2$address))
```

We then create a *get_coords* function below to obtain the x,y coordinates of the addresses:

```{r}
get_coords <- function(add_list){
  
  # Create a data frame to store all retrieved coordinates
  postal_coords <- data.frame()
    
  for (i in add_list){
    #print(i)

    r <- GET('https://www.onemap.gov.sg/api/common/elastic/search?',
           query=list(searchVal=i,
                     returnGeom='Y',
                     getAddrDetails='Y'))
    data <- fromJSON(rawToChar(r$content))
    found <- data$found
    res <- data$results
    
    # Create a new data frame for each address
    new_row <- data.frame()
    
    # If single result, append 
    if (found == 1){
      postal <- res$POSTAL 
      lat <- res$LATITUDE
      lng <- res$LONGITUDE
      new_row <- data.frame(address= i, 
                            postal = postal, 
                            latitude = lat, 
                            longitude = lng)
    }
    
    # If multiple results, drop NIL and append top 1
    else if (found > 1){
      # Remove those with NIL as postal
      res_sub <- res[res$POSTAL != "NIL", ]
      
      # Set as NA first if no Postal
      if (nrow(res_sub) == 0) {
          new_row <- data.frame(address= i, 
                                postal = NA, 
                                latitude = NA, 
                                longitude = NA)
      }
      
      else{
        top1 <- head(res_sub, n = 1)
        postal <- top1$POSTAL 
        lat <- top1$LATITUDE
        lng <- top1$LONGITUDE
        new_row <- data.frame(address= i, 
                              postal = postal, 
                              latitude = lat, 
                              longitude = lng)
      }
    }

    else {
      new_row <- data.frame(address= i, 
                            postal = NA, 
                            latitude = NA, 
                            longitude = NA)
    }
    
    # Add the row
    postal_coords <- rbind(postal_coords, new_row)
  }
  return(postal_coords)
}
```

We obtain the coordinates from the two lists using the code chunk below:

```{r}
#| eval: false
coords1 <- get_coords(add_list1)
coords2 <- get_coords(add_list2)
```

We then join both data frames together:

```{r}
#| eval: false
coords <- bind_rows(coords1, coords2) %>%
  distinct(postal, .keep_all = TRUE)
```

We save the coords as a new rds file and load it into the R environment to avoid re-running the above steps:

```{r}
#| eval: false
write_rds(coords, "data/rds/coords.rds")
```

```{r}
coords <- read_rds("data/rds/coords.rds")
```

### 5.1.3 Adding the coords data to resale data frame and removing unwanted variables

We then add the x, y coordinates data to the *resale* data frame while also just keeping variables that would be needed for further analysis:

```{r}
resale_tidy <- left_join(resale, coords, by = "address") %>%
  select(month, storey_range, floor_area_sqm, flat_model, resale_price, remaining_lease_mth, age, latitude, longitude)
```

### 5.1.4 Recoding "storey_range" and "flat_model" variables

We note that the variables "storey_range" and "flat_model" are categorical variables specifically:

-   "storey_range" is in character format "XX TO YY" i.e. "01 TO 03" up till "40 TO 42"

-   "flat_model" is in character format i.e. "Terrace", "DBSS"

To faciliate downstream regression analysis, we will need to recode these categorical variables. We carry out label encoding by first converting the variables into factors and subsequently as integers:

::: panel-tabset
## "storey_range"

```{r}
ggplot(resale_tidy,
       aes(x = storey_range))+
  geom_bar(fill = "skyblue")+
  ggtitle("Distribution of storey range")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
resale_tidy$storey_range <-  as.integer(factor(resale_tidy$storey_range))
```

We note that this labels "01 TO 03" as 1, so on and so forth such that "40 TO 42" is labelled as 14.

## "flat_model"

```{r}
ggplot(resale_tidy,
       aes(x = flat_model))+
  geom_bar(fill = "lightyellow")+
  ggtitle("Distribution of flat model")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
resale_tidy$flat_model <-  as.integer(factor(resale_tidy$flat_model))
```

We note that this labels the flat model types as follows:

| Flat Model        | Label |
|-------------------|-------|
| DBSS              | 1     |
| Improved          | 2     |
| Model A           | 3     |
| New Generation    | 4     |
| Premium Apartment | 5     |
| Simplified        | 6     |
| Standard          | 7     |
| Terrace           | 8     |
:::

### 5.1.5 Jittering

We convert the *resale_tidy* data frame into a simple features data object and apply `st_transform()` to change the EPSG code to 3414:

```{r}
resale_tidy <- st_as_sf(resale_tidy, coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 3414)
```

We then apply `st_jitter()` as HDB units can share the same postal code:

```{r}
resale_tidy <- resale_tidy %>%
  st_jitter(amount = .1)
```

### 5.1.5 Saving as a new rds file

We save the output as a new rds file and load it into the R environment:

```{r}
#| eval: false
write_rds(resale_tidy,"data/rds/resale_tidy.rds")
```

```{r}
resale_tidy <- read_rds("data/rds/resale_tidy.rds")
```

### 5.2 Locational Factors Data Wrangling

As seen in [3. The Data], we will utilise several locational data for this take-home exercise - some with geographic coordinates and some without. For the latter, we will need to run a similar reverse geocoding step as that in [5.1.2 Reverse geocoding using One Map API] to obtain the x, y coordinates of the locational factors.

### 5.2.1 Locational factors with geographic coordinates

We will first import the locational factors data into the R environment:

-   Note that for this group of factors, we will also apply the `st_transform()` function to ensure that the EPSG code is 3414.

    ::: panel-tabset
    ## Eldercare

    ```{r}
    eldercare <- st_read(dsn = "data/geospatial",
                         layer = "ELDERCARE") %>%
      st_transform(3414)
    ```

    We note that the *eldercare* sf has a point geometry.

    ## Hawker Centre

    ```{r}
    hawker <- st_read("data/geospatial/HawkerCentresGEOJSON.geojson") %>%
      st_transform(crs = 3414)
    ```

    We note that the *hawker* sf object has a point geometry.

    ## Parks

    ```{r}
    parks <- st_read("data/geospatial/Parks.geojson") %>%
      st_transform(crs = 3414)
    ```

    We note that the *parks* sf object has a point geometry.

    ## Supermarkets

    ```{r}
    supermarkets <- st_read("data/geospatial/SupermarketsGEOJSON.geojson") %>%
      st_transform(crs = 3414)
    ```

    We note that the *supermarkets* sf object has a point geometry.

    ## CHAS Clinics

    ```{r}
    chas <- st_read("data/geospatial/CHASClinics.geojson")%>%
      st_transform(crs = 3414)
    ```

    We note that the *chas* sf object has a point geometry.

    ## Childcare Services

    ```{r}
    childcare <- st_read("data/geospatial/ChildCareServices.geojson") %>%
      st_transform(crs = 3414)
    ```

    We note that the *childcare* sf object has a point geometry.

    ## Kindergartens

    ```{r}
    kindergartens <- st_read("data/geospatial/Kindergartens.geojson") %>%
      st_transform(crs = 3414)
    ```

    We note that the *kindergartens* sf object has a point geometry.

    ## MRT

    ```{r}
    mrt <- st_read(dsn = "data/geospatial", 
                   layer = "RapidTransitSystemStation") %>%
      st_transform(crs = 4326)
    ```

    We note that the *mrt* sf object has a polygon geometry and that a warning message is shown in the output above, indicating that there are some invalid geometry within the sf object i.e. Non closed ring detected. However, error occurred after we tried to apply `st_make_valid()` function.

    Based on the code chunk below, we note that if we had tried to retain just rows of mrt stations with valid geometry, we will miss out on 3 observations as the total number of observations would drop from 230 to 227. Upon examination, we note that the 3 rows was for "UPPER THOMSON MRT STATION", "HARBOURFRONT MRT STATION" and "BOCC".

    ```{r}
    mrt_clean <- mrt %>%
      filter(st_is_valid(.)) %>%  
      filter(!is.na(geometry))
    ```

    Given that having "UPPER THOMSON MRT STATION" and "HARBOURFRONT MRT STATION" would be important to retain in the mrt data, we utilise another method to derive the coordinates of all mrt stations. We first extract the names of the mrt stations in the *mrt* sf object:

    ```{r}
    mrt_names <- mrt %>%
      select(STN_NAM_DE) %>%
      st_drop_geometry()
    ```

    We then extract unique mrt names into the list - we note that there are only 203 unique mrt names out of the 230 observations:

    ```{r}
    add_list <- sort(unique(mrt_names$STN_NAM_DE))
    ```

    We then utilise the get_coords function to obtain the x, y coordinates of the mrt stations:

    ```{r}
    coords_mrt_names <- get_coords(add_list)
    ```

    We save the output as a new rds file for safekeeping:

    ```{r}
    write_rds(coords_mrt_names,"data/rds/coords_mrt_names.rds")
    ```

    ```{r}
    coords_mrt_names <- read_rds("data/rds/coords_mrt_names.rds")
    ```

    We then utilise `left_join()` to join the coordinates data with *mrt_names:*

    ```{r}
    mrt_names <- left_join(mrt_names,coords_mrt_names, by = c("STN_NAM_DE" = "address"))
    ```

    However based on the code chunk below, we note that the postal code for "HUME MRT STATION" and "FOUNDERS' MEMORIAL MRT STATION" were NA hence leading to NA latitude and longitude.

    ```{r}
    rows_with_na <- mrt_names %>%
      filter(if_any(everything(), is.na))
    rows_with_na
    ```

    This is understandable as these two stations are upcoming stations and their postal codes might not be accurately retrieved. We manually obtain the postal codes via Google search:

    -   Hume MRT Station: 589681

    -   Founders' Memorial MRT Station: 438965

    We then create a list to apply the get_coords() function on:

    ```{r}
    #| eval: false
    addedmrt <- list(589681,438965)
    coords_addedmrt <- get_coords(addedmrt)
    ```

    ```{r}
    mrt_names <- mrt_names %>%
      mutate(
        postal = case_when(
          STN_NAM_DE == "HUME MRT STATION" ~ "589681",
          STN_NAM_DE == "FOUNDERS' MEMORIAL MRT STATION" ~ "438965",
          TRUE ~ as.character(postal)  # Default to existing value
        ),
        latitude = case_when(
          STN_NAM_DE == "HUME MRT STATION" ~ "1.35502264316802",
          STN_NAM_DE == "FOUNDERS' MEMORIAL MRT STATION" ~ "1.29141358573109",
          TRUE ~ as.character(latitude)  # Default to existing value
        ),
        longitude = case_when(
          STN_NAM_DE == "HUME MRT STATION" ~ "103.768680287307",
          STN_NAM_DE == "FOUNDERS' MEMORIAL MRT STATION" ~ "103.868706430815",
          TRUE ~ as.character(longitude)  # Default to existing value
        )
      )
    ```

    We double check if there are na rows using the code chunk below:

    ```{r}
    rows_with_na <- mrt_names %>%
      filter(if_any(everything(), is.na))
    rows_with_na
    ```

    We note that there are no longer any missing values.

    We convert *mrt_names* as a sf object, apply `st_transform()` to ensure that it is in the SVY21 projected coordinate system (EPSG 3414) and save it as a new rds file:

    ```{r}
    #| eval: false
    mrt_names <- st_as_sf(mrt_names, coords = c("longitude", "latitude"), crs = 4326) %>%   
      st_transform(crs = 3414)
    ```

    ```{r}
    #| eval: false
    write_rds(mrt_names,"data/rds/mrt_names.rds")
    ```

    ```{r}
    mrt_names <- read_rds("data/rds/mrt_names.rds")
    mrt_names
    ```

    We note that the mrt data has a point geometry.

    ## Bus Stops

    ```{r}
    busstop <- st_read("data/geospatial",
    layer = "BusStop") %>%
      st_transform(crs = 3414)
    ```

    We note that the *busstops* sf object has a point geometry.
    :::

    ### 5.2.2 Locational factors without geographic coordinates

::: panel-tabset
## Primary school

To obtain primary school data, we download the csv file on ["General Information of schools"](https://data.gov.sg/collections/457/view) from Data.gov.sg. We then filter the data to primary and mixed levels schools, noting that there are schools that comprise both primary and secondary levels under the same name. In this step, we filter out schools that are a mix of secondary and tertiary levels i.e. Temasek Junior College.

Further, we retain the postal code field for the purpose of carrying out reverse geocoding in the subsequent step. Note that in this case we have also converted the postal codes for three primary schools, namely Cantonment Primary School, CHIJ Kellock and Radin Mas Primary School as an initial trial of drawing the x, y coordinates from the postal codes yielded NA x,y coordinates for these three schools. This was due to the missing zero in front of the postal codes in the original dataset.

```{r}
allschools <- read_csv("data/aspatial/schools.csv") 
```

```{r}
prisch <- allschools %>%
  filter(mainlevel_code %in% c("PRIMARY","MIXED LEVELS")) %>%
  filter(!school_name %in% c("TEMASEK JUNIOR COLLEGE","SINGAPORE SPORTS SCHOOL","SCHOOL OF THE ARTS, SINGAPORE","RIVER VALLEY HIGH SCHOOL
","RAFFLES INSTITUTION","NUS HIGH SCHOOL OF MATHEMATICS AND SCIENCE
","NATIONAL JUNIOR COLLEGE
","HWA CHONG INSTITUTION
","ANGLO-CHINESE SCHOOL (INDEPENDENT)")) %>%
  select(school_name,postal_code) %>%
  mutate(postal_code = case_when(
    postal_code == "88256" ~ "088256",
    postal_code == "99757" ~ "099757",
    postal_code == "99840" ~ "099840",
    TRUE ~ as.character(postal_code)
  ))
```

We then generate a unique list of postal codes from the *schools* data frame:

```{r}
add_list <- sort(unique(prisch$postal_code))
```

We then apply the *get_coords* function to obtain the x, y coordinates and save it as a new rds file to avoid re-running the process:

```{r}
#| eval: false
coords_sch <- get_coords(add_list)
```

```{r}
#| eval: false
write_rds(coords_sch, "data/rds/coords_sch.rds")
```

```{r}
coords_sch <- read_rds("data/rds/coords_sch.rds")
```

We then `left_join()` the *schools* data frame with *coords_sch:*

```{r}
#| eval: false
prisch <- left_join(prisch, coords_sch, by = c("postal_code" = "postal")) %>%
  select(c(-address))
```

We convert the *prisch* data frame into a simple features data object and apply `st_transform()` to change the EPSG code to 3414:

```{r}
#| eval: false
prisch <- st_as_sf(prisch, coords = c("longitude", "latitude"), crs = 4326) %>%   
  st_transform(crs = 3414)
```

We save the output as a new rds file:

```{r}
#| eval: false
write_rds(prisch, "data/rds/prisch.rds")
```

```{r}
prisch <- read_rds("data/rds/prisch.rds")
```

## Good primary school

The [Salary.sg](https://www.salary.sg/2024/best-primary-schools-2024-by-popularity/) forum ranks all 180 primary schools in Singapore. To study how the proximity of good primary schools to HDB will affect resale prices, we will extract and focus on the top 20% primary schools (36 out of 180) for this locational factor:

```{r}
goodprisch <- read_csv("data/aspatial/goodprisch.csv") %>%
  mutate(school = toupper(school)) %>%
  .[1:36,]
```

As there are only the names of the primary school, we will need to obtain their x, y coordinates by doing a `left_join` with *prisch* data frame that was earlier imported.

However, before we do so, we note that there could be a mix of curly and straight apostrophes in school names which could create issues when we do a `left_join()`. We hence standardize the apostrophes to a straight one using the code chunk below:

```{r}
goodprisch <- goodprisch %>%
  mutate(school = trimws(gsub("['’]", "'", school)))

prisch <- prisch %>%
  mutate(school_name = trimws(gsub("['’]", "'", school_name)))
```

We then do a `left_join()` however we manually input the postal codes for four schools for which the namings slightly differ between both data frames:

```{r}
goodprisch <- left_join(goodprisch, prisch, by = c("school" = "school_name")) %>%
  select(school,postal_code) %>%
  mutate(postal_code = case_when(
    school == "CHIJ ST. NICHOLAS GIRLS'" ~ "569405",
    school == "SINGAPORE CHINESE GIRLS" ~ "309437",
    school == "KUO CHUAN PRESBYTERIAN PRIMARY" ~ "579793",
    school == "PEI HWA PRESBYTERIAN PRIMARY" ~ "597610",
    TRUE ~ as.character(postal_code)
  ),postal_code = ifelse(is.na(postal_code), "309437", postal_code))
```

::: callout-note
Note that the last line of code above was added as there was difficulty in indicating that the postal code for SINGAPORE CHINESE GIRLS was 309437. As it was the only NA entry after the postal codes had been adjusted, a last line was used to manually make the postal code for the school 309437.
:::

We then generate a unique list of postal codes from the *goodprisch* data frame:

```{r}
add_list <- sort(unique(goodprisch$postal_code))
```

We then apply the *get_coords* function to obtain the x, y coordinates and save it as a new rds file to avoid re-running the process:

```{r}
#| eval: false
coords_goodprisch <- get_coords(add_list)
```

```{r}
#| eval: false
write_rds(coords_goodprisch, "data/rds/coords_goodprisch.rds")
```

```{r}
coords_goodprisch <- read_rds("data/rds/coords_goodprisch.rds")
```

We then `left_join()` the *goodprisch* data frame with *coords_goodprisch:*

```{r}
#| eval: false
goodprisch <- left_join(goodprisch, coords_goodprisch, by = c("postal_code" = "postal")) %>%
  select(c(-address))
```

We convert the *goodprisch* data frame into a simple features data object and apply `st_transform()` to change the EPSG code to 3414:

```{r}
#| eval: false
goodprisch <- st_as_sf(goodprisch, coords = c("longitude", "latitude"), crs = 4326) %>%   
  st_transform(crs = 3414)
```

We save the output as a new rds file:

```{r}
#| eval: false
write_rds(goodprisch, "data/rds/goodprisch.rds")
```

```{r}
goodprisch <- read_rds("data/rds/goodprisch.rds")
```

## CBD

We obtain the coordinates of Singapore's CBD via Google. Based on this [site](https://www.findlatitudeandlongitude.com/l/central+business+district%2C+singapore/424476/), the latitude and longitude are 1.352083, 103.819836:

```{r}
#| eval: false
latitude <- 1.352083
longitude <- 103.819836

cbd_coords <- data.frame(
  longitude = longitude,
  latitude = latitude
)

cbd <- st_as_sf(cbd_coords, coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs= 3414)
```

We save this as a new rds file:

```{r}
#| eval: false
write_rds(cbd, "data/rds/cbd.rds")
```

```{r}
cbd <- read_rds("data/rds/cbd.rds")
```

## Shopping malls

A list of shopping malls was generated with reference to [Wikipedia](https://en.wikipedia.org/wiki/List_of_shopping_malls_in_Singapore):

```{r}
shoppingmall <- read_csv("data/aspatial/shopping_mall.csv")
```

To ensure that there are no duplicates, we generate a unique list using the code chunk below:

```{r}
add_list <- sort(unique(shoppingmall$shopping_mall))
```

As we only have the shopping mall names, we need to apply the *get_coords* function to obtain the x, y coordinates and save it as a new rds file to avoid re-running the process:

```{r}
#| eval: false
coords_mall <- get_coords(add_list)
```

```{r}
#| eval: false
write_rds(coords_mall, "data/rds/coords_mall.rds")
```

```{r}
coords_mall <- read_rds("data/rds/coords_mall.rds")
```

We then `left_join()` the *shoppingmall* data frame with *coords_mall:*

```{r}
#| eval: false
shoppingmall <- left_join(shoppingmall, coords_mall, by = c("shopping_mall" = "address"))
```

We convert the *shoppingmall* data frame into a simple features data object and apply `st_transform()` to change the EPSG code to 3414:

```{r}
#| eval: false
shoppingmall <- st_as_sf(shoppingmall, coords = c("longitude", "latitude"), crs = 4326) %>%   
  st_transform(crs = 3414)
```

We save the output as a new rds file:

```{r}
#| eval: false
write_rds(shoppingmall, "data/rds/shoppingmall.rds")
```

```{r}
shoppingmall <- read_rds("data/rds/shoppingmall.rds")
```
:::

### 5.2.3 Generating proximity variables using geoprocessing functions

From the locational factors above, we will generate proximity variables that will be used for downstream geographically weighted regression analysis, specifically we will utilise geoprocessing functions:

-   `st_distance()` function to create variables representing the proximity to the nearest eldercare, hawker centre, park, supermarket, CHAS clinic, childcare service, kindergarten, MRT, bus stop, primary school (in general), good primary school (based on most popular 20%), CBD and shopping mall

-   `st_buffer()`, `st_intersects()` and `length()` to create a buffer and carry out point-in-polygon counts to determine:

    1.  Number of hawker centres within 350m
    2.  Number of supermarkets within 350m
    3.  Number of childcare centres within 350m
    4.  Number of kindergartens within 350m
    5.  Number of bus stops within 350m
    6.  Number of primary schools within 1km

    These six variables were chosen as they represent essential services that most might need on a day-to-day basis i.e. for meals, grocery shopping, education and transport.

    350m was set for the first five variables as the [average habitual gait of adults in Singapore is 1.08 metres/second](https://pmc.ncbi.nlm.nih.gov/articles/PMC11205154/) which makes 350m an approximately 5 minutes walking distance, a good representation of how close/convenient day-to-day facilities are in the neighbourhood which could impact HDB resale prices.

    For the sixth variable, the distance was set at 1km as [priority admission to primary schools is given to those living within 1km of the school](https://www.moe.gov.sg/primary/p1-registration/distance).

#### 5.2.3.1 Creating proximity variables using `st_distance()`

We utilise `st_distance()` to calculate the distance matrix between each row in *resale_tidy* and all rows in the different locational factors listed below and then utilise `apply()` to find the minimum (nearest) distance for each row of HDB resale data to the nearest locational factor:

::: panel-tabset
## Eldercare

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_ELDERCARE = st_distance(., eldercare) %>% 
                                   apply(1, min))
```

## Hawker Centre

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_HAWKER = st_distance(., hawker) %>% 
                                   apply(1, min))
```

## Park

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_PARK = st_distance(., parks) %>% 
                                   apply(1, min))
```

## Supermarket

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_SUPERMARKET = st_distance(., supermarkets) %>% 
                                   apply(1, min))
```

## CHAS clinic

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_CLINIC = st_distance(., chas) %>% 
                                   apply(1, min))
```

## Childcare service

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_CHILDCARE= st_distance(., childcare) %>% 
                                   apply(1, min))
```

## Kindergarten

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_KINDERGARTEN = st_distance(., kindergartens) %>% 
                                   apply(1, min))
```

## MRT

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_MRT = st_distance(., mrt_names) %>% 
                                   apply(1, min))
```

## Bus stop

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_BUSSTOP = st_distance(., busstop) %>% 
                                   apply(1, min))
```

## Primary school (in general)

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_PRISCH = st_distance(., prisch) %>% 
                                   apply(1, min))
```

## Good primary school

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_GOODPRISCH = st_distance(., goodprisch) %>% 
                                   apply(1, min))
```

## CBD

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_CBD = st_distance(., cbd) %>% 
                                   apply(1, min))
```

## Shopping mall

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_MALL = st_distance(., shoppingmall) %>% 
                                   apply(1, min))
```
:::

We save the updated *resale_tidy* as a new rds data file for safekeeping:

```{r}
#| eval: false
write_rds(resale_tidy, "data/rds/resale_tidy_stdist.rds")
```

```{r}
resale_tidy_stdist <- read_rds("data/rds/resale_tidy_stdist.rds")
```

#### 5.2.3.2 Creating proximity variables using `st_buffer()`, `st_intersects()` and `length()`

We will first create two data frames which comprise a buffer of 350 metres and 1km around each HDB unit in the *resale_tidy* sf object:

```{r}
buffer_350m <- st_buffer(resale_tidy,
                         dist = 350)
buffer_1km <- st_buffer(resale_tidy,
                        dist = 1000)
```

We then utilise `st_intersects()` and `lengths()` to count the number of essential services within 350m and 1km of the HDB resale units in *resale_tidy:*

::: panel-tabset
## No. of hawker centres within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_HAWKER <- lengths(st_intersects(buffer_350m, hawker))
```

## No. of supermarkets within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_SUPERMARKETS <- lengths(st_intersects(buffer_350m, supermarkets))
```

## No. of childcare centres within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_CHILDCARE <- lengths(st_intersects(buffer_350m, childcare))
```

## No. of kindergartens within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_KINDERGARTENS <- lengths(st_intersects(buffer_350m, kindergartens))
```

## No. of bus stops within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_BUSSTOP <- lengths(st_intersects(buffer_350m, busstop))
```

## No. of primary schools within 1km

```{r}
#| eval: false
buffer_1km$WITHIN_1KM_PRISCH <- lengths(st_intersects(buffer_1km, prisch))
```
:::

We save *buffer_350m* and *buffer_1km* as new rds files for safekeeping:

```{r}
#| eval: false
write_rds(buffer_350m,"data/rds/buffer_350m.rds")
write_rds(buffer_1km,"data/rds/buffer_1km.rds")
```

```{r}
buffer_350m <- read_rds("data/rds/buffer_350m.rds")
buffer_1km <- read_rds("data/rds/buffer_1km.rds")
```

We then combine the WITHIN_1KM_PRISCH field from *buffer_1km* to *buffer_350m* to put all the variables together:

```{r}
#| eval: false
buffer_350m <- buffer_350m %>%
  st_drop_geometry()
buffer_1km <- buffer_1km %>%
  st_drop_geometry()
resale_tidy_all <- cbind(resale_tidy, buffer_350m[,c(21:25)])
resale_tidy_all <- cbind(resale_tidy_all, buffer_1km[,21])
```

We save *resale_tidy_all* as a new rds file and load it into the R environment for further analysis:

```{r}
#| eval: false
write_rds(resale_tidy_all,"data/rds/resale_tidy_all.rds")
```

```{r}
resale_tidy_all <- read_rds("data/rds/resale_tidy_all.rds")
```

### 5.3 Geospatial Data Wrangling

We will load the Master Plan 2019 Subzone Boundary data using the code chunk below, note that we also apply `st_transform()` to ensure that the data file has the right EPSG code 3414 which is the projected coordinate system for Singapore:

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

We check if the EPSG code has been correctly transformed using the code chunk below:

```{r}
st_crs(mpsz)
```

We note that the *mpsz* has a multipolygon geometry. We save the output as a new rds file for further analysis:

```{r}
#| eval: false
write_rds(mpsz,"data/rds/mpsz.rds")
```

```{r}
mpsz <- read_rds("data/rds/mpsz.rds")
```

## 5.4 Computing Correlation Matrix

Before loading the independent variables/predictors into the predictive model, we will need to use the correlation matrix to examine for signs of multicollinearity. This helps to avoid biasness/skewness in the predictive models that may occur if highly correlated variables were used.

```{r}
resale_tidy_all_nogeo <- resale_tidy_all %>%
  st_drop_geometry()

corrplot(cor(resale_tidy_all_nogeo[,c(2:4,6:26)]),
         diag = FALSE,
         order = "AOE",
         tl.pos = "td",
         tl.cex = 0.5,
         number.cex = 0.7,
         method = "number",
         type = "upper")
```

Based on the correlation matrix, we there is multicollinearity between variables "age" and "remaining_lease_mth" which are highly negatively correlated (-1.00). As "age" is less specific than "remaining_lease_mth", given that we used the current year 2024 to minus the lease commencement year and did not consider month in its formulation, we will retain the latter for further analysis:

```{r}
resale_tidy_all <- resale_tidy_all %>%
  select(-age)
```

## 5.5 Data Sampling

Before we carry out further analysis, we will split the *resale_tidy_all* data into train, validation and test data sets. As the intent of the exercise is to predict the HDB resale prices between July and September 2024, we will extract the data during this period as the test data set. For the remaining data, we will use `initial_split()` of the [**rsample**](https://rsample.tidymodels.org/)package to split it into train (65%) and validation (35%):

```{r}
#| eval: false 
set.seed(1234)
test_data <- resale_tidy_all %>%
  filter(month >= "2024-07")
remaining_data <- resale_tidy_all %>%
  filter(month < "2024-07")

resale_split <- initial_split(remaining_data,prop = 6.5/10,)
train_data <- training(resale_split) 
val_data <- testing(resale_split)
```

We save the train, validation and test data sets as new rds files and load them into the R environment to avoid re-running the step above:

```{r}
#| eval: false 
write_rds(train_data,"data/rds/train_data.rds")
write_rds(val_data,"data/rds/val_data.rds")
write_rds(test_data,"data/rds/test_data.rds")
```

```{r}
train_data <- read_rds("data/rds/train_data.rds")
val_data <- read_rds("data/rds/val_data.rds")
test_data <- read_rds("data/rds/test_data.rds")
```

## 5.6 Multiple linear regression (MLR)

### 5.6.1 Non-spatial MLR

```{r}
#| eval: false
price_mlr <- lm(resale_price ~ floor_area_sqm + storey_order + remaining_lease_mths + PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER + PROX_MRT + PROX_PARK+ PROX_MALL + PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN + WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
                data = train_data)
```

### 5.6.2 Geographically weighted MLR predictive model

-   Locational factors

    -   Use st_distance() or function in sf package

        -   Proximity to CBD

        -   Proximity to eldercare

        -   Proximity to foodcourt/hawker centres

        -   Proximity to MRT

        -   Proximity to park

        -   Proximity to good primary school

        -   Proximity to shopping mall

        -   Proximity to supermarket

    -   Generate a buffer from HDB and use point-in-geometry

        -   Numbers of kindergartens within 350m

        -   Numbers of childcare centres within 350m

        -   Numbers of bus stop within 350m

        -   Numbers of primary school within 1km

-   Structural factors (done)

    -   "floor_area_sqm": Area of the unit

    -   "storey_range": Floor level

    -   "remaining_lease_mth": Remaining lease

    -   "age": Age of the unit
