---
title: "Take-home Exercise 3"
author: "Kock Si Min"
date: October 26, 2024
date-modified: "last-modified"
toc: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

# **Take-home Exercise 3: Predicting HDB Resale Prices with Geographically Weighted Machine Learning Methods**

## 1. Setting the Scene

Housing is a sign of household wealth. It is a major investment for most people and its price is affected by several factors that can be global in nature such as the general economy of a country, inflation rate or are factors specific to the nature of the property which can be further divided into structural and locational factors:

-   [Structural variables]{.underline} are related to the property such as the size, fitting, tenure

-   [Locational variables]{.underline} are related to the neighbourhood of the property such as proximity to childcare services, public transport and shopping centres.

Conventionally, predictive models for housing resale prices were built by using the Ordinary Least Square (OLS) method. However, this method failed to take into consideration that spatial autocorrelation and spatial heterogeneity exist in geographic data sets such as housing transactions. With the existence of spatial autocorrelation, the OLS estimation of predictive housing resale pricing models could lead to biased, inconsistent, or inefficient results (Anselin 1998). In view of this limitation, **Geographical Weighted Models** were introduced to better calibrate predictive models for housing resale prices.

## 2. The Task

In this take-home exercise, we will calibrate predictive models to predict HDB resale prices between July-September 2024 by using HDB resale transaction records in 2023. Specifically, we will develop models using **conventional multiple linear regression**, **random forest** and **geographically weighted random forest**, and evaluate the performance of the models in predicting HDB resale prices.

Our study will be scoped to focus on 3-room flat types, given the [increasing resale price of 3-room HDB flats in recent years](https://www.asiaone.com/money/more-3-room-resale-flats-change-hands-800k-and-higher-2024).

In this exercise, we will predict HDB resale prices, the dependent variable, with independent variables listed below:

-   [**Structural factors:**]{.underline} Storey range, Floor area, Flat model, Remaining lease

-   [**Locational factors**]{.underline} which have been broadly categorised as:

    -   **Essential services:**

        -   Proximity to childcare services, kindergartens, primary schools, good primary schools, eldercare services, hawker centres, supermarkets, clinics, MRT and bus stops

    -   **Leisure:**

        -   Proximity to parks, malls

Besides proximity, we will also zoom in on the number of essential services within a certain parameter of the HDB resale units:

-   Number of hawker centres, supermarkets, childcare services, kindergartens, bus stops within 350m

-   Number of primary schools within 1km

These six variables were chosen as they represent essential services that most might need on a day-to-day basis i.e. for meals, grocery shopping, education and transport.

350m was set for the first five variables as the [average habitual gait of adults in Singapore is 1.08 metres/second](https://pmc.ncbi.nlm.nih.gov/articles/PMC11205154/) which makes 350m an approximately 5 minutes walking distance, a good representation of how close/convenient day-to-day facilities are in the neighbourhood which could impact HDB resale prices.

For the last variable, the distance was set at 1km as [priority admission to primary schools is given to those living within 1km of the school](https://www.moe.gov.sg/primary/p1-registration/distance).

## 3. The Data

The following datasets will be used:

-   **Aspatial Data:**

    -   [HDB Resale Flat Prices](https://data.gov.sg/datasets/d_8b84c4ee58e3cfc0ece0d773c8ca6abc/view) from [Data.gov.sg](https://data.gov.sg/): A list of HDB resale transacted prices in Singapore from Jan 2017 onwards. As mentioned, our focus will be on 3-room flat types, given the [increasing resale price of 3-room HDB flats in recent years.](https://www.asiaone.com/money/more-3-room-resale-flats-change-hands-800k-and-higher-2024)

-   **Geospatial dataset**:

    -   URA's Master Plan 2019 Subzone Boundary from [Data.gov.sg](https://data.gov.sg/)

    -   Locational factors with geographic coordinates from [Data.gov.sg](https://data.gov.sg/)

        | S/N | Data                                |
        |-----|-------------------------------------|
        | 1   | **Eldercare** services in Singapore |
        | 2   | **Hawker Centres** in Singapore     |
        | 3   | **Parks** in Singapore              |
        | 4   | **Supermarkets** in Singapore       |
        | 5   | **CHAS clinics** in Singapore       |
        | 6   | **Childcare services** in Singapore |
        | 7   | **Kindergartens** in Singapore      |

    -   Locational factors with geographic coordinates from [Datamall.lta.gov.sg](https://datamall.lta.gov.sg/content/datamall/en.html):

        | S/N | Data                                                                |
        |-----|---------------------------------------------------------------------|
        | 1   | **MRT/LRT stations** in Singapore with the station names and codes. |
        | 2   | **Bus stops** in Singapore                                          |

    -   Locational factors without geographic coordinates from [Data.gov.sg](https://data.gov.sg/) and other sources:

        | S/N | Data                                                                  | Source                                                                                                                |
        |-----|-----------------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------|
        | 1   | **Primary school data**                                               | [Data.gov.sg](https://data.gov.sg/) - extracted from the list on General information of schools                       |
        | 2   | **Good primary schools** ordered in terms of popularity for year 2024 | [Local Salary Forum](https://www.salary.sg/2024/best-primary-schools-2024-by-popularity/) - extracted into a csv file |
        | 3   | **CBD** coordinates                                                   | Google                                                                                                                |
        | 4   | **Shopping malls** in Singapore                                       | [Wikipedia](https://en.wikipedia.org/wiki/List_of_shopping_malls_in_Singapore)                                        |

## 4. Installing and Loading R packages

We will install and load the following R packages:

```{r}
pacman::p_load(tidyverse, sf, GWmodel, SpatialML, tmap, rsample, Metrics, MLmetrics,httr, jsonlite, rvest, fastDummies, corrplot, olsrr,DT,ggpubr,kableExtra,performance,spatstat,spdep)
```

The rationale for key packages are explained below:

-   [**tidyverse**](https://cran.r-project.org/web/packages/tidyverse/index.html): to import, wrangle and visualise data

-   [**sf**](https://cran.r-project.org/web/packages/sf/index.html)**:** to import, manage and process geospatial data

-   [**GWmodel**](https://cran.r-project.org/web/packages/GWmodel/)**:** to calibrate geographical weighted family of models

-   [**SpatialML**](https://cran.r-project.org/web/packages/SpatialML/index.html)**:** to perform random forest regressions

-   [**tmap**](https://cran.r-project.org/web/packages/tmap/)**:** provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [leaflet](https://leafletjs.com/) API

-   [**httr**](https://httr.r-lib.org/): to allow R to talk to http

-   [**rvest**](https://rvest.tidyverse.org/): used to crawl websites

-   [**jsonlite**](https://cran.r-project.org/web/packages/jsonlite/): to communicate with one map API and convert data into data frame format

-   [**spatstat**](https://spatstat.org/)**:** useful functions for point pattern analysis

-   [**spdep**](https://cran.r-project.org/web/packages/spdep/index.html): to compute spatial weights, global and local spatial autocorrelation statistics

-   [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)**, [ggpubr](https://rpkgs.datanovia.com/ggpubr/)** and [**performance**](https://easystats.github.io/performance/): for multivariate data visualisation and analysis

-   [**olsrr**](https://olsrr.rsquaredacademy.com/)**:** to build OLS and perform diagnostics tests

-   [**Metrics**](https://cran.r-project.org/web/packages/Metrics/Metrics.pdf) and [**MLmetrics**](https://cran.r-project.org/web/packages/MLmetrics/MLmetrics.pdf): to compute evaluation metrics for machine learning models

## 5. Aspatial Data Wrangling

### 5.1 HDB Resale Prices Data Wrangling

We will load the HDB Resale Flat Prices data using `read_csv()` function in the code chunk below and we filter the data to

\(i\) retain data from 2023 onwards as the intent of the take-home exercise is to predict HDB resale prices in July to September 2024 using 2023 data and

\(ii\) focus on just 3-room flat types.

```{r}
resale <- read_csv("data/aspatial/hdbresale.csv") %>%
  filter(month >= "2023-01" & month <= "2024-09" & flat_type == "3 ROOM")
```

A glimpse at the data indicates that there are 11642 observations across 11 variables including the dependent variable in our study "resale_price":

```{r}
glimpse(resale)
```

### 5.1.1 Creating new variables

Based on the output above, we also note that the address of the HDB flats have been split into "block", "street_name" and that there is no postal code. To facilitate the development of geographically weighted predictive models, we will need to combine the "block" and "street_name" to form a new field "address" and utilise [One Map API](https://www.onemap.gov.sg/apidocs/) to carry out reverse geocoding i.e. carry out API calls to return x, y coordinates for the address that is passed through the system.

In the code chunk below, we will create "address" as a new field and as we do so, we also split the "remaining_lease" field which is currently in the "XX years and YY months" format to "remaining_lease_yr" and "remaining_lease_mth" respectively. We then convert the "remaining_lease_yr" to months by multiplying it by 12 and add it to the "remaining_lease_mth" field to obtain the total remaining lease in terms of months:

```{r}
resale <- resale %>%
  mutate(address = paste(block, street_name)) %>%
  mutate(remaining_lease_yr = as.integer(str_extract(remaining_lease, "^[0-9]+"))) %>%
  mutate(remaining_lease_mth = as.integer(str_extract(remaining_lease, "(?<=years, )[0-9]+"))) %>%
  mutate(remaining_lease_mth = ifelse(is.na(remaining_lease_mth), 0, remaining_lease_mth)) %>%
  mutate(remaining_lease_mth = remaining_lease_yr * 12 + remaining_lease_mth) %>%
  mutate(age = 2024 - lease_commence_date)
```

We double check that the above code was run properly using the code chunk below that checks for NA values:

```{r}
na_rows <- resale[rowSums(is.na(resale)) > 0, ]
na_rows
```

We note that there are no NA values.

### 5.1.2 Reverse geocoding using One Map API

To carry reverse geocoding, we will need to create a list of unique addresses from the *resale* dataset to avoid duplicative API calls to the same address. However, it is important to note that we should not push too much data i.e. 10,000 to the system as it will result in a longer lag time before results can be returned.

We hence first split the *resale* data so that we divide the 11642 observations into two:

```{r}
resale1 <- resale %>%
  filter(month <= "2023-10")

resale2 <- resale %>%
  filter(month > "2023-10" & month <= "2024-09")
```

We utilise the *resale1* and *resale2* datasets to generate two lists of unique addresses:

```{r}
add_list1 <- sort(unique(resale1$address))
add_list2 <- sort(unique(resale2$address))
```

We then create a *get_coords* function below to obtain the x,y coordinates of the addresses:

```{r}
get_coords <- function(add_list){
  
  # Create a data frame to store all retrieved coordinates
  postal_coords <- data.frame()
    
  for (i in add_list){
    #print(i)

    r <- GET('https://www.onemap.gov.sg/api/common/elastic/search?',
           query=list(searchVal=i,
                     returnGeom='Y',
                     getAddrDetails='Y'))
    data <- fromJSON(rawToChar(r$content))
    found <- data$found
    res <- data$results
    
    # Create a new data frame for each address
    new_row <- data.frame()
    
    # If single result, append 
    if (found == 1){
      postal <- res$POSTAL 
      lat <- res$LATITUDE
      lng <- res$LONGITUDE
      new_row <- data.frame(address= i, 
                            postal = postal, 
                            latitude = lat, 
                            longitude = lng)
    }
    
    # If multiple results, drop NIL and append top 1
    else if (found > 1){
      # Remove those with NIL as postal
      res_sub <- res[res$POSTAL != "NIL", ]
      
      # Set as NA first if no Postal
      if (nrow(res_sub) == 0) {
          new_row <- data.frame(address= i, 
                                postal = NA, 
                                latitude = NA, 
                                longitude = NA)
      }
      
      else{
        top1 <- head(res_sub, n = 1)
        postal <- top1$POSTAL 
        lat <- top1$LATITUDE
        lng <- top1$LONGITUDE
        new_row <- data.frame(address= i, 
                              postal = postal, 
                              latitude = lat, 
                              longitude = lng)
      }
    }

    else {
      new_row <- data.frame(address= i, 
                            postal = NA, 
                            latitude = NA, 
                            longitude = NA)
    }
    
    # Add the row
    postal_coords <- rbind(postal_coords, new_row)
  }
  return(postal_coords)
}
```

We obtain the coordinates from the two lists using the code chunk below:

```{r}
#| eval: false
coords1 <- get_coords(add_list1)
coords2 <- get_coords(add_list2)
```

We then join both data frames together to obtain a consolidated data frame of coordinates:

```{r}
#| eval: false
coords <- bind_rows(coords1, coords2) %>%
  distinct(postal, .keep_all = TRUE)
```

We save the coords as a new rds file and load it into the R environment to avoid re-running the above steps:

```{r}
#| eval: false
write_rds(coords, "data/rds/coords.rds")
```

```{r}
coords <- read_rds("data/rds/coords.rds")
```

### 5.1.3 Adding the coords data to resale data frame and removing unwanted variables

We then add the x, y coordinates data to the *resale* data frame via a `left_join()` while also just keeping variables that would be needed for further analysis using the `select()` function:

```{r}
resale_tidy <- left_join(resale, coords, by = "address") %>%
  select(month, storey_range, floor_area_sqm, flat_model, resale_price, remaining_lease_mth, age, latitude, longitude)
```

### 5.1.4 Recoding "storey_range" and "flat_model" variables

We note that the variables "storey_range" and "flat_model" are categorical variables specifically:

-   "storey_range" is in character format "XX TO YY" i.e. "01 TO 03" up till "40 TO 42"

-   "flat_model" is in character format i.e. "Terrace", "DBSS"

To faciliate downstream regression analysis, we will need to recode these categorical variables. We carry out label encoding by first converting the variables into factors and subsequently as integers:

::: panel-tabset
## "storey_range"

```{r}
#| code-fold: true
ggplot(resale_tidy,
       aes(x = storey_range))+
  geom_bar(fill = "skyblue")+
  ggtitle("Distribution of storey range")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
resale_tidy$storey_range <-  as.integer(factor(resale_tidy$storey_range))
```

We note that this labels "01 TO 03" as 1, so on and so forth such that "40 TO 42" is labelled as 14.

## "flat_model"

```{r}
#| code-fold: true
ggplot(resale_tidy,
       aes(x = flat_model))+
  geom_bar(fill = "lightyellow")+
  ggtitle("Distribution of flat model")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
resale_tidy$flat_model <-  as.integer(factor(resale_tidy$flat_model))
```

We note that this labels the flat model types as follows:

| Flat Model        | Label |
|-------------------|-------|
| DBSS              | 1     |
| Improved          | 2     |
| Model A           | 3     |
| New Generation    | 4     |
| Premium Apartment | 5     |
| Simplified        | 6     |
| Standard          | 7     |
| Terrace           | 8     |
:::

### 5.1.5 Saving as a sf object

We convert the *resale_tidy* data frame into a simple features data object and apply `st_transform()` to change the EPSG code to 3414:

```{r}
resale_tidy <- st_as_sf(resale_tidy, coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 3414)
```

### 5.1.6 Saving as a new rds file

We save the output as a new rds file and load it into the R environment:

```{r}
#| eval: false
write_rds(resale_tidy,"data/rds/resale_tidy.rds")
```

```{r}
resale_tidy <- read_rds("data/rds/resale_tidy.rds")
```

### 5.2 Locational Factors Data Wrangling

As seen in [3. The Data], we will utilise several locational data for this take-home exercise - some with geographic coordinates and some without. For the latter, we will need to run a similar reverse geocoding step as that in [5.1.2 Reverse geocoding using One Map API] to obtain the x, y coordinates of the locational factors.

### 5.2.1 Locational factors with geographic coordinates

We will first import the locational factors data into the R environment using the `st_read()` function.

Note that for this group of factors, we will also apply the `st_transform()` function to ensure that the EPSG code is 3414.

::: panel-tabset
## Eldercare

```{r}
eldercare <- st_read(dsn = "data/geospatial",
                     layer = "ELDERCARE") %>%
  st_transform(3414)
```

We note that the *eldercare* sf has a point geometry.

## Hawker Centre

```{r}
hawker <- st_read("data/geospatial/HawkerCentresGEOJSON.geojson") %>%
  st_transform(crs = 3414)
```

We note that the *hawker* sf object has a point geometry.

## Parks

```{r}
parks <- st_read("data/geospatial/Parks.geojson") %>%
  st_transform(crs = 3414)
```

We note that the *parks* sf object has a point geometry.

## Supermarkets

```{r}
supermarkets <- st_read("data/geospatial/SupermarketsGEOJSON.geojson") %>%
  st_transform(crs = 3414)
```

We note that the *supermarkets* sf object has a point geometry.

## CHAS Clinics

```{r}
chas <- st_read("data/geospatial/CHASClinics.geojson")%>%
  st_transform(crs = 3414)
```

We note that the *chas* sf object has a point geometry.

## Childcare Services

```{r}
childcare <- st_read("data/geospatial/ChildCareServices.geojson") %>%
  st_transform(crs = 3414)
```

We note that the *childcare* sf object has a point geometry.

## Kindergartens

```{r}
kindergartens <- st_read("data/geospatial/Kindergartens.geojson") %>%
  st_transform(crs = 3414)
```

We note that the *kindergartens* sf object has a point geometry.

## MRT

```{r}
mrt <- st_read(dsn = "data/geospatial", 
               layer = "RapidTransitSystemStation") %>%
  st_transform(crs = 4326)
```

We note that the *mrt* sf object has a polygon geometry and that a warning message appears when we run the code above, indicating that there are some invalid geometry within the sf object i.e. Non closed ring detected. Error would still occur even as we tried to apply `st_make_valid()` function.

Based on the code chunk below, we note that if we had tried to retain just rows of mrt stations with valid geometry, we will miss out on three observations as the total number of observations would drop from 230 to 227.

```{r}
mrt_clean <- mrt %>%
  filter(st_is_valid(.)) %>%  
  filter(!is.na(geometry))

mrt_clean
```

Upon examination, we note that the 3 affected rows were for "UPPER THOMSON MRT STATION", "HARBOURFRONT MRT STATION" and "BOCC".

Given that "UPPER THOMSON MRT STATION" and "HARBOURFRONT MRT STATION" would be important to retain in the mrt data, we utilise another method to derive the coordinates of all mrt stations. We first extract the names of the mrt stations in the *mrt* sf object:

```{r}
mrt_names <- mrt %>%
  select(STN_NAM_DE) %>%
  st_drop_geometry()
```

We then extract unique mrt names into the list - we note that there are only 203 unique mrt names out of the 230 observations:

```{r}
add_list <- sort(unique(mrt_names$STN_NAM_DE))
```

We then utilise the get_coords function to carry out API calls to obtain the x, y coordinates of the mrt stations:

```{r}
coords_mrt_names <- get_coords(add_list)
```

We save the output as a new rds file for safekeeping:

```{r}
write_rds(coords_mrt_names,"data/rds/coords_mrt_names.rds")
```

```{r}
coords_mrt_names <- read_rds("data/rds/coords_mrt_names.rds")
```

We then utilise `left_join()` to join the coordinates data with *mrt_names:*

```{r}
mrt_names <- left_join(mrt_names,coords_mrt_names, by = c("STN_NAM_DE" = "address"))
```

However based on the code chunk below, we note that the postal code for "HUME MRT STATION" and "FOUNDERS' MEMORIAL MRT STATION" were NA hence leading to NA latitude and longitude.

```{r}
rows_with_na <- mrt_names %>%
  filter(if_any(everything(), is.na))
rows_with_na
```

This is understandable as these two stations are upcoming stations and their postal codes might not be accurately retrieved. We manually obtain the postal codes via Google search:

-   Hume MRT Station: 589681

-   Founders' Memorial MRT Station: 438965

We then create a list to apply the get_coords() function on:

```{r}
#| eval: false
addedmrt <- list(589681,438965)
coords_addedmrt <- get_coords(addedmrt)
```

We then incorporate the two coordinates into the *mrt_names* data frame*:*

```{r}
mrt_names <- mrt_names %>%
  mutate(
    postal = case_when(
      STN_NAM_DE == "HUME MRT STATION" ~ "589681",
      STN_NAM_DE == "FOUNDERS' MEMORIAL MRT STATION" ~ "438965",
      TRUE ~ as.character(postal)  # Default to existing value
    ),
    latitude = case_when(
      STN_NAM_DE == "HUME MRT STATION" ~ "1.35502264316802",
      STN_NAM_DE == "FOUNDERS' MEMORIAL MRT STATION" ~ "1.29141358573109",
      TRUE ~ as.character(latitude)  # Default to existing value
    ),
    longitude = case_when(
      STN_NAM_DE == "HUME MRT STATION" ~ "103.768680287307",
      STN_NAM_DE == "FOUNDERS' MEMORIAL MRT STATION" ~ "103.868706430815",
      TRUE ~ as.character(longitude)  # Default to existing value
    )
  )
```

We double check if there are na rows using the code chunk below:

```{r}
rows_with_na <- mrt_names %>%
  filter(if_any(everything(), is.na))
rows_with_na
```

We note that there are no longer any missing values.

We convert *mrt_names* to a sf object, apply `st_transform()` to ensure that it is in the SVY21 projected coordinate system (EPSG 3414) and save it as a new rds file:

```{r}
#| eval: false
mrt_names <- st_as_sf(mrt_names, coords = c("longitude", "latitude"), crs = 4326) %>%   
  st_transform(crs = 3414)
```

```{r}
#| eval: false
write_rds(mrt_names,"data/rds/mrt_names.rds")
```

```{r}
mrt_names <- read_rds("data/rds/mrt_names.rds")
mrt_names
```

We note that the mrt data has a point geometry.

## Bus Stops

```{r}
busstop <- st_read("data/geospatial",
layer = "BusStop") %>%
  st_transform(crs = 3414)
```

We note that the *busstops* sf object has a point geometry.
:::

### 5.2.2 Locational factors without geographic coordinates

::: panel-tabset
## Primary school

To obtain primary school data, we download the csv file on ["General Information of schools"](https://data.gov.sg/collections/457/view) from Data.gov.sg. We then filtered the data to primary and mixed levels schools, noting that there are schools that comprise both primary and secondary levels under the same name. In this step, we filter out schools that are a mix of secondary and tertiary levels i.e. Temasek Junior College.

Further, we retain the postal code field for the purpose of carrying out reverse geocoding in the subsequent step. Note that in this case we have also converted the postal codes for three primary schools, namely Cantonment Primary School, CHIJ Kellock and Radin Mas Primary School as an initial trial of drawing the x, y coordinates from the postal codes yielded NA x,y coordinates for these three schools. This was due to the missing zero in front of the postal codes in the original dataset.

```{r}
allschools <- read_csv("data/aspatial/schools.csv") 
```

```{r}
prisch <- allschools %>%
  filter(mainlevel_code %in% c("PRIMARY","MIXED LEVELS")) %>%
  filter(!school_name %in% c("TEMASEK JUNIOR COLLEGE","SINGAPORE SPORTS SCHOOL","SCHOOL OF THE ARTS, SINGAPORE","RIVER VALLEY HIGH SCHOOL
","RAFFLES INSTITUTION","NUS HIGH SCHOOL OF MATHEMATICS AND SCIENCE
","NATIONAL JUNIOR COLLEGE
","HWA CHONG INSTITUTION
","ANGLO-CHINESE SCHOOL (INDEPENDENT)")) %>%
  select(school_name,postal_code) %>%
  mutate(postal_code = case_when(
    postal_code == "88256" ~ "088256",
    postal_code == "99757" ~ "099757",
    postal_code == "99840" ~ "099840",
    TRUE ~ as.character(postal_code)
  ))
```

We then generate a unique list of postal codes from the *schools* data frame:

```{r}
add_list <- sort(unique(prisch$postal_code))
```

We then apply the *get_coords* function to carry out the API calls to obtain the x, y coordinates and save it as a new rds file to avoid re-running the process:

```{r}
#| eval: false
coords_sch <- get_coords(add_list)
```

```{r}
#| eval: false
write_rds(coords_sch, "data/rds/coords_sch.rds")
```

```{r}
coords_sch <- read_rds("data/rds/coords_sch.rds")
```

We then `left_join()` the *schools* data frame with *coords_sch:*

```{r}
#| eval: false
prisch <- left_join(prisch, coords_sch, by = c("postal_code" = "postal")) %>%
  select(c(-address))
```

We convert the *prisch* data frame into a simple features data object and apply `st_transform()` to change the EPSG code to 3414:

```{r}
#| eval: false
prisch <- st_as_sf(prisch, coords = c("longitude", "latitude"), crs = 4326) %>%   
  st_transform(crs = 3414)
```

We save the output as a new rds file:

```{r}
#| eval: false
write_rds(prisch, "data/rds/prisch.rds")
```

```{r}
prisch <- read_rds("data/rds/prisch.rds")
```

## Good primary school

The [Salary.sg](https://www.salary.sg/2024/best-primary-schools-2024-by-popularity/) forum ranks all 180 primary schools in Singapore. To study how the proximity of good primary schools to HDB will affect resale prices, we will extract and focus on the top 20% primary schools (36 out of 180) for this locational factor:

```{r}
goodprisch <- read_csv("data/aspatial/goodprisch.csv") %>%
  mutate(school = toupper(school)) %>%
  .[1:36,]
```

As there are only the names of the primary school, we will need to obtain their x, y coordinates by doing a `left_join` with *prisch* data frame that was earlier imported.

However, before we do so, we note that there could be a mix of curly and straight apostrophes in school names which could create issues when we do a `left_join()`. We hence standardize the apostrophes to a straight one using the code chunk below:

```{r}
goodprisch <- goodprisch %>%
  mutate(school = trimws(gsub("['’]", "'", school)))

prisch <- prisch %>%
  mutate(school_name = trimws(gsub("['’]", "'", school_name)))
```

We then do a `left_join()` however we manually input the postal codes for four schools as their namings slightly differ between both data frames:

```{r}
goodprisch <- left_join(goodprisch, prisch, by = c("school" = "school_name")) %>%
  select(school,postal_code) %>%
  mutate(postal_code = case_when(
    school == "CHIJ ST. NICHOLAS GIRLS'" ~ "569405",
    school == "SINGAPORE CHINESE GIRLS" ~ "309437",
    school == "KUO CHUAN PRESBYTERIAN PRIMARY" ~ "579793",
    school == "PEI HWA PRESBYTERIAN PRIMARY" ~ "597610",
    TRUE ~ as.character(postal_code)
  ),postal_code = ifelse(is.na(postal_code), "309437", postal_code))
```

::: callout-note
Note that the last line of code above was added as there was difficulty in indicating that the postal code for SINGAPORE CHINESE GIRLS was 309437. As it was the only NA entry after the postal codes had been adjusted, a last line was used to manually make the postal code for the school 309437.
:::

We then generate a unique list of postal codes from the *goodprisch* data frame:

```{r}
add_list <- sort(unique(goodprisch$postal_code))
```

We then apply the *get_coords* function to obtain the x, y coordinates and save it as a new rds file to avoid re-running the process:

```{r}
#| eval: false
coords_goodprisch <- get_coords(add_list)
```

```{r}
#| eval: false
write_rds(coords_goodprisch, "data/rds/coords_goodprisch.rds")
```

```{r}
coords_goodprisch <- read_rds("data/rds/coords_goodprisch.rds")
```

We then `left_join()` the *goodprisch* data frame with *coords_goodprisch:*

```{r}
#| eval: false
goodprisch <- left_join(goodprisch, coords_goodprisch, by = c("postal_code" = "postal")) %>%
  select(c(-address))
```

We convert the *goodprisch* data frame into a simple features data object and apply `st_transform()` to change the EPSG code to 3414:

```{r}
#| eval: false
goodprisch <- st_as_sf(goodprisch, coords = c("longitude", "latitude"), crs = 4326) %>%   
  st_transform(crs = 3414)
```

We save the output as a new rds file:

```{r}
#| eval: false
write_rds(goodprisch, "data/rds/goodprisch.rds")
```

```{r}
goodprisch <- read_rds("data/rds/goodprisch.rds")
```

## CBD

We obtain the coordinates of Singapore's CBD via Google. Based on Google, we found this [site](https://www.findlatitudeandlongitude.com/l/central+business+district%2C+singapore/424476/) which state the latitude and longitude of the CBD to be 1.352083, 103.819836. We first create a data frame *cbd_coords* then convert it to a sf object and apply `st_transform()` to ensure that the EPSG code would be 3414, so that as with all other geographical data, the data is in the SVY21 projected coordinates system.

```{r}
#| eval: false
latitude <- 1.352083
longitude <- 103.819836

cbd_coords <- data.frame(
  longitude = longitude,
  latitude = latitude
)

cbd <- st_as_sf(cbd_coords, coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs= 3414)
```

We save this as a new rds file:

```{r}
#| eval: false
write_rds(cbd, "data/rds/cbd.rds")
```

```{r}
cbd <- read_rds("data/rds/cbd.rds")
```

## Shopping malls

A list of shopping malls was obtained from [Wikipedia](https://en.wikipedia.org/wiki/List_of_shopping_malls_in_Singapore) and extracted into a csv file which is loaded using `read_csv()` in the code chunk below:

```{r}
shoppingmall <- read_csv("data/aspatial/shopping_mall.csv")
```

To ensure that there are no duplicates, we generate a unique list using the code chunk below:

```{r}
add_list <- sort(unique(shoppingmall$shopping_mall))
```

As we only have the shopping mall names, we need to apply the *get_coords* function to carry out API calls to obtain the x, y coordinates. We save it as a new rds file to avoid re-running the process:

```{r}
#| eval: false
coords_mall <- get_coords(add_list)
```

```{r}
#| eval: false
write_rds(coords_mall, "data/rds/coords_mall.rds")
```

```{r}
coords_mall <- read_rds("data/rds/coords_mall.rds")
```

We then `left_join()` the *shoppingmall* data frame with *coords_mall:*

```{r}
#| eval: false
shoppingmall <- left_join(shoppingmall, coords_mall, by = c("shopping_mall" = "address"))
```

We convert the *shoppingmall* data frame into a simple features data object and apply `st_transform()` to change the EPSG code to 3414:

```{r}
#| eval: false
shoppingmall <- st_as_sf(shoppingmall, coords = c("longitude", "latitude"), crs = 4326) %>%   
  st_transform(crs = 3414)
```

We save the output as a new rds file:

```{r}
#| eval: false
write_rds(shoppingmall, "data/rds/shoppingmall.rds")
```

```{r}
shoppingmall <- read_rds("data/rds/shoppingmall.rds")
```
:::

### 5.2.3 Generating proximity variables using geoprocessing functions

From the locational factors above, we will generate proximity variables that will be used for downstream geographically weighted regression analysis, specifically we will utilise geoprocessing functions:

-   `st_distance()` function to create variables representing the proximity to the nearest eldercare, hawker centre, park, supermarket, CHAS clinic, childcare service, kindergarten, MRT, bus stop, primary school (in general), good primary school (based on most popular 20%), CBD and shopping mall

-   `st_buffer()`, `st_intersects()` and `length()` to create a buffer and carry out point-in-polygon counts to determine:

    1.  Number of hawker centres within 350m
    2.  Number of supermarkets within 350m
    3.  Number of childcare centres within 350m
    4.  Number of kindergartens within 350m
    5.  Number of bus stops within 350m
    6.  Number of primary schools within 1km

    The rationale for these factors have been explained in section [2. The Task].

#### 5.2.3.1 Creating proximity variables using `st_distance()`

We utilise `st_distance()` to calculate the distance matrix between each row in *resale_tidy* and all rows in the different locational factors listed below and then utilise `apply()` to find the minimum (nearest) distance for each row of HDB resale data to the nearest locational factor:

::: panel-tabset
## Eldercare

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_ELDERCARE = st_distance(., eldercare) %>% 
                                   apply(1, min))
```

## Hawker Centre

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_HAWKER = st_distance(., hawker) %>% 
                                   apply(1, min))
```

## Park

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_PARK = st_distance(., parks) %>% 
                                   apply(1, min))
```

## Supermarket

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_SUPERMARKET = st_distance(., supermarkets) %>% 
                                   apply(1, min))
```

## CHAS clinic

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_CLINIC = st_distance(., chas) %>% 
                                   apply(1, min))
```

## Childcare service

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_CHILDCARE= st_distance(., childcare) %>% 
                                   apply(1, min))
```

## Kindergarten

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_KINDERGARTEN = st_distance(., kindergartens) %>% 
                                   apply(1, min))
```

## MRT

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_MRT = st_distance(., mrt_names) %>% 
                                   apply(1, min))
```

## Bus stop

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_BUSSTOP = st_distance(., busstop) %>% 
                                   apply(1, min))
```

## Primary school (in general)

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_PRISCH = st_distance(., prisch) %>% 
                                   apply(1, min))
```

## Good primary school

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_GOODPRISCH = st_distance(., goodprisch) %>% 
                                   apply(1, min))
```

## CBD

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_CBD = st_distance(., cbd) %>% 
                                   apply(1, min))
```

## Shopping mall

```{r}
#| eval: false
resale_tidy <- resale_tidy %>%
  mutate(PROX_MALL = st_distance(., shoppingmall) %>% 
                                   apply(1, min))
```
:::

We save the updated *resale_tidy* as a new rds data file for safekeeping:

```{r}
#| eval: false
write_rds(resale_tidy, "data/rds/resale_tidy_stdist.rds")
```

```{r}
resale_tidy <- read_rds("data/rds/resale_tidy_stdist.rds")
```

#### 5.2.3.2 Creating proximity variables using `st_buffer()`, `st_intersects()` and `length()`

We will first create two data frames which comprise a buffer of 350 metres and 1km around each HDB unit in the *resale_tidy* sf object using `st_buffer()` in the code chunk below:

```{r}
buffer_350m <- st_buffer(resale_tidy,
                         dist = 350)
buffer_1km <- st_buffer(resale_tidy,
                        dist = 1000)
```

We then utilise `st_intersects()` and `lengths()` to count the number of essential services within 350m and 1km of the HDB resale units in *resale_tidy:*

::: panel-tabset
## No. of hawker centres within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_HAWKER <- lengths(st_intersects(buffer_350m, hawker))
```

## No. of supermarkets within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_SUPERMARKETS <- lengths(st_intersects(buffer_350m, supermarkets))
```

## No. of childcare centres within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_CHILDCARE <- lengths(st_intersects(buffer_350m, childcare))
```

## No. of kindergartens within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_KINDERGARTENS <- lengths(st_intersects(buffer_350m, kindergartens))
```

## No. of bus stops within 350m

```{r}
#| eval: false
buffer_350m$WITHIN_350M_BUSSTOP <- lengths(st_intersects(buffer_350m, busstop))
```

## No. of primary schools within 1km

```{r}
#| eval: false
buffer_1km$WITHIN_1KM_PRISCH <- lengths(st_intersects(buffer_1km, prisch))
```
:::

We save *buffer_350m* and *buffer_1km* as new rds files for safekeeping:

```{r}
#| eval: false
write_rds(buffer_350m,"data/rds/buffer_350m.rds")
write_rds(buffer_1km,"data/rds/buffer_1km.rds")
```

```{r}
buffer_350m <- read_rds("data/rds/buffer_350m.rds")
buffer_1km <- read_rds("data/rds/buffer_1km.rds")
```

We then combine the WITHIN_1KM_PRISCH field from *buffer_1km* to *buffer_350m* to put all the variables together:

```{r}
#| eval: false
buffer_350m <- buffer_350m %>%
  st_drop_geometry()
buffer_1km <- buffer_1km %>%
  st_drop_geometry()
resale_tidy_all <- cbind(resale_tidy, buffer_350m[,c(21:25)])
resale_tidy_all <- cbind(resale_tidy_all, buffer_1km[,21])
```

We save *resale_tidy_all* as a new rds file and load it into the R environment for further analysis:

```{r}
#| eval: false
write_rds(resale_tidy_all,"data/rds/resale_tidy_all.rds")
```

```{r}
resale_tidy_all <- read_rds("data/rds/resale_tidy_all.rds")
```

### 5.3 Geospatial Data Wrangling

We will load the Master Plan 2019 Subzone Boundary data using the code chunk below, note that we also apply `st_transform()` to ensure that the data file has the right EPSG code 3414 which is the projected coordinate system for Singapore:

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

We check if the EPSG code has been correctly transformed using the code chunk below:

```{r}
st_crs(mpsz)
```

We note that the *mpsz* has a multipolygon geometry. We save the output as a new rds file for further analysis:

```{r}
#| eval: false
write_rds(mpsz,"data/rds/mpsz.rds")
```

```{r}
mpsz <- read_rds("data/rds/mpsz.rds")
```

### 5.4 Exploratory Data Analysis (EDA)

We will perform some EDA to better understand the data.

### 5.4.1 Distribution of HDB resale price

We study the distribution of the resale price of 3-room HDB flats using the code chunk below:

```{r}
#| code-fold: true
ggplot(resale_tidy_all,
       aes(x = `resale_price`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "turquoise")+
  labs(title = "Distribution of 3-room HDB resale price", x = "HDB Resale Price")
```

### 5.4.2 Distribution of structural and locational factors

We will also visualise the distribution of the various structural and locational factors by creating individual plots for each variable and then using `ggarrange()` to organise the plots into 3 column by 8 row multiple small plots:

```{r}
#| fig-height: 15
#| fig-width: 10
#| code-fold: true
storeyrange <- ggplot(resale_tidy_all,
                      aes(x=`storey_range`))+
  geom_bar(color = "black",
           fill = "violetred")

floorareasqm <- ggplot(resale_tidy_all,
                      aes(x=`floor_area_sqm`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "turquoise")

flatmodel <- ggplot(resale_tidy_all,
                      aes(x=`floor_area_sqm`))+
  geom_bar(color = "black",
           fill = "tomato")

remaininglease <- ggplot(resale_tidy_all,
                      aes(x=`remaining_lease_mth`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "thistle2")

PROXELDER <- ggplot(resale_tidy_all,
                      aes(x=`PROX_ELDERCARE`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "steelblue1")

PROXHAWKER <- ggplot(resale_tidy_all,
                      aes(x=`PROX_HAWKER`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "slategray2")

PROXPARK <- ggplot(resale_tidy_all,
                      aes(x=`PROX_PARK`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "skyblue")

PROXSUPERMARKET <- ggplot(resale_tidy_all,
                      aes(x=`PROX_SUPERMARKET`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "seashell3")

PROXCLINIC <- ggplot(resale_tidy_all,
                      aes(x=`PROX_CLINIC`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "seagreen2")

PROXCHILDCARE <- ggplot(resale_tidy_all,
                      aes(x=`PROX_CHILDCARE`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "salmon")

PROXKINDERGARTEN <- ggplot(resale_tidy_all,
                      aes(x=`PROX_KINDERGARTEN`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "royalblue")

PROXMRT <- ggplot(resale_tidy_all,
                      aes(x=`PROX_MRT`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "rosybrown2")

PROXBUS <- ggplot(resale_tidy_all,
                      aes(x=`PROX_BUSSTOP`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "plum")

PROXPRISCH <- ggplot(resale_tidy_all,
                      aes(x=`PROX_PRISCH`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "pink1")

PROXGOODPRISCH <- ggplot(resale_tidy_all,
                      aes(x=`PROX_GOODPRISCH`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "palevioletred2")

PROXCBD <- ggplot(resale_tidy_all,
                      aes(x=`PROX_CBD`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "paleturquoise2")

PROXMALL <- ggplot(resale_tidy_all,
                      aes(x=`PROX_MALL`))+
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "palegreen2")

within350mhawker <- ggplot(resale_tidy_all,
                      aes(x=`WITHIN_350M_HAWKER`))+
  geom_bar(color = "black",
           fill = "orchid1")

within350msupmarket <- ggplot(resale_tidy_all,
                      aes(x=`WITHIN_350M_SUPERMARKETS`))+
  geom_bar(color = "black",
           fill = "olivedrab2")

within350mcc <- ggplot(resale_tidy_all,
                      aes(x=`WITHIN_350M_CHILDCARE`))+
  geom_bar(color = "black",
           fill = "mistyrose2")

within350mkinder <- ggplot(resale_tidy_all,
                      aes(x=`WITHIN_350M_KINDERGARTENS`))+
  geom_bar(color = "black",
           fill = "lightpink2")

within350mbusstop <- ggplot(resale_tidy_all,
                      aes(x=`WITHIN_350M_BUSSTOP`))+
  geom_bar(color = "black",
           fill = "lightgoldenrod1")

within1kmprisch <- ggplot(resale_tidy_all,
                      aes(x=`WITHIN_1KM_PRISCH`))+
  geom_bar(color = "black",
           fill = "lavender")

ggarrange(storeyrange, floorareasqm, flatmodel, remaininglease, PROXELDER, PROXHAWKER, PROXPARK, PROXSUPERMARKET, PROXCLINIC, PROXCHILDCARE, PROXKINDERGARTEN, PROXMRT, PROXBUS, PROXPRISCH,  PROXGOODPRISCH, PROXCBD,  PROXMALL, within350mhawker, within350msupmarket, within350mcc, within350mkinder, within350mbusstop, within1kmprisch,
         ncol = 3,
         nrow = 8)
```

### 5.4.3 Geographical Distribution of 3-room HDB Resale Price

We will also study the geospatial distribution of the 3-room HDB resale prices using the tmap package. Note that we have the first line in the code chunk below to prevent errors in the plotting of our high quality cartographic map. We also set the zoom limits to be between the minimum and maximum zoom level of 11 and 14 respectively.

```{r}
#| code-fold: true
tmap_options(check.and.fix = TRUE)

tmap_mode("view")
tm_shape(mpsz)+
  tm_polygons()+
  tm_shape(resale_tidy_all)+
  tm_dots(col = "resale_price",
          alpha = 0.6,
          style = "quantile")+
  tm_view(set.zoom.limits = c(11,14))
```

From the map above, we note that there is a concentration of higher-priced 3-room HDB resale flat in the northeast region of Singapore. We also note that 3-room HDB resale flats in the west are usually of a lower resale price.

Before we move on to next steps, we must remember to change the tmap mode to plot using the code chunk below:

```{r}
tmap_mode("plot")
```

### 5.5 Computing Correlation Matrix

Before loading the independent variables/predictors into the predictive model, we will need to use the correlation matrix to examine for signs of multicollinearity. This helps to avoid biasness/skewness in the predictive models that may occur if highly correlated variables were used.

In the code chunk below, we will drop the geometry from *resale_tidy_all* before using `corrplot()` function to generate the correlation matrix.

```{r}
resale_tidy_all_nogeo <- resale_tidy_all %>%
  st_drop_geometry()

corrplot(cor(resale_tidy_all_nogeo[,c(2:4,6:26)]),
         diag = FALSE,
         order = "AOE",
         tl.pos = "td",
         tl.cex = 0.5,
         number.cex = 0.7,
         method = "number",
         type = "upper")
```

::: {.callout-note style="color: steelblue"}
Based on the correlation matrix, we observe that there is multicollinearity between variables "age" and "remaining_lease_mth" which are highly negatively correlated (-1.00).

As "age" is less specific than "remaining_lease_mth", given that we used the current year 2024 to minus the lease commencement year and did not consider month in its formulation, we will retain the latter for further analysis:
:::

We drop the "age" variable in the code chunk below:

```{r}
resale_tidy_all <- resale_tidy_all %>%
  select(-age)
```

### 5.6 Jittering and Data Sampling

Before we carry out further analysis, we will

\(i\) apply jittering as several HDB units share the same postal code and this will cause downstream issues for our predictive model analysis. This is proven as we check for duplicated points in our *resale_tidy_all* sf object using the code chunk below:

```{r}
any(duplicated(resale_tidy_all))
```

The output of the above is TRUE indicating that there are duplicated points as expected. We hence apply jittering below, note that the amount follows the units of coordinates. For this case, the units of measurement for EPSG 3414 is in metres. The code chunk below hence jitters points using 100 metres.

```{r}
resale_tidy_all <- st_jitter(resale_tidy_all, amount = 100)
```

We confirm that jittering has been carried out properly using the code chunk below:

```{r}
any(duplicated(resale_tidy_all))
```

\(ii\) split the *resale_tidy_all* data into train and test data sets. As the intent of the exercise is to predict the HDB resale prices between July and September 2024, we will extract the data during this period as the test data set. We will use the remaining data as train data.

```{r}
#| eval: false 
test_data <- resale_tidy_all %>%
  filter(month >= "2024-07")
train_data <- resale_tidy_all %>%
  filter(month < "2024-07")
```

We save the train, validation and test data sets as new rds files and load them into the R environment to avoid re-running the step above:

```{r}
#| eval: false 
write_rds(train_data,"data/rds/train_data.rds")
write_rds(test_data,"data/rds/test_data.rds")
```

```{r}
train_data <- read_rds("data/rds/train_data.rds")
test_data <- read_rds("data/rds/test_data.rds")
```

### 5.7 Multiple linear regression (MLR)

We start off by building a conventional non-spatial MLR model using the `lm()` function in the code chunk below:

```{r}
#| eval: false
price_mlr <- lm(resale_price ~ storey_range + floor_area_sqm + flat_model + remaining_lease_mth + PROX_ELDERCARE + PROX_HAWKER + PROX_PARK + PROX_SUPERMARKET + PROX_CLINIC + PROX_CHILDCARE + PROX_KINDERGARTEN + PROX_MRT + PROX_BUSSTOP + PROX_PRISCH + PROX_GOODPRISCH + PROX_CBD + PROX_MALL + WITHIN_350M_HAWKER + WITHIN_350M_SUPERMARKETS + WITHIN_350M_CHILDCARE + WITHIN_350M_KINDERGARTENS +  WITHIN_350M_BUSSTOP + WITHIN_1KM_PRISCH,
                data = train_data)
```

We save the output as a new rds file and load it into the R environment to avoid re-running the above step:

```{r}
#| eval: false
write_rds(price_mlr,"data/rds/price_mlr.rds")
```

```{r}
price_mlr <- read_rds("data/rds/price_mlr.rds")
```

We generate the report using the `olsrr_regress()` function:

```{r}
ols_regress(price_mlr)
```

::: {.callout-note style="color: steelblue"}
Based on the report output above that the adjusted R2 is 0.678 which indicate that the predictors are able to explain 67.8% of the HDB resale prices.

However, at 95% confidence level, we note that not all the independent variables are statistically significant. Specifically, PROX_BUSSTOP, PROX_PRISCH, PROX_MALL and WITHIN_350M_BUSSTOP are not statistically significant.
:::

We revise the model by removing these variables and calibrate the revised model using the code chunk below:

```{r}
#| eval: false
price_mlr1 <- lm(resale_price ~ storey_range + floor_area_sqm + flat_model + remaining_lease_mth + PROX_ELDERCARE + PROX_HAWKER + PROX_PARK + PROX_SUPERMARKET + PROX_CLINIC + PROX_CHILDCARE + PROX_KINDERGARTEN + PROX_MRT + PROX_GOODPRISCH + PROX_CBD + WITHIN_350M_HAWKER + WITHIN_350M_SUPERMARKETS + WITHIN_350M_CHILDCARE + WITHIN_350M_KINDERGARTENS +  WITHIN_1KM_PRISCH,
                data = train_data)
```

As with the above, we save the output as a new rds file and load it into the R environment to avoid re-running the above step:

```{r}
#| eval: false
write_rds(price_mlr1,"data/rds/price_mlr1.rds")
```

```{r}
price_mlr1 <- read_rds("data/rds/price_mlr1.rds")
```

We generate the report using the `olsrr_regress()` function:

```{r}
ols_regress(price_mlr1)
```

::: {.callout-note style="color: steelblue"}
We note that all remaining independent variables are significant, however we note that the adjusted R2 still remained at 0.678, which indicates that this revised model is only able to predict as well as the earlier model and explain approximately 67.8% of 3-room HDB resale prices.
:::

### 5.7.1 Test for multicollinearity

We also check once more for multicollinearity by analysing the variance inflation factors (VIF) using the \``check_collinearity()` function in the performance package. We then use `kable()` to generate a table from the tibble data frame created:

```{r}
vif <- check_collinearity(price_mlr1) 

kable(vif,       caption = "Variance Inflation Factor (VIF) results") %>%   kable_styling(font_size = 18)
```

We also plot the vif values generated for easier visualisation:

```{r}
plot(vif)+
  theme(axis.text.x = element_text(angle = 45, hjust =1))
```

::: {.callout-note style="color: steelblue"}
Based on the output above, we note that the VIF values are all less than 5, indicating that there's no presence of multicollinearity.
:::

### 5.7.2 Test for non-linearity

In MLR, it is also important to test for the linearity and additivity of the relationship between dependent and independent variables. This is because MLR assumes a linear relationship between each predictor and the outcome variable. If a predictor has a non-linear relationship with the dependent variable, it can lead to biased or inaccurate predictions.

We hence utilise the `ols_plot_resid_fit()` function of the olsrr package to perform the linearity assumption test, to test the validity of our model:

```{r}
ols_plot_resid_fit(price_mlr1)
```

::: {.callout-note style="color: steelblue"}
The plot above indicates that most of the data points are scattered around the 0 line, hence we can conclude that the relationships between the dependent variable and the independent variables are linear.
:::

### 5.7.3 Test for Normality Assumption

For MLR, we would also need to assess the normality of residuals. Normality of residuals mean that they are distributed symmetrically around zero, with no skewness, implying that the model would be able to capture the main patterns and sources of variation in the data and that errors are random and independent ([source](https://www.linkedin.com/advice/0/what-best-methods-assess-normality-residuals#:~:text=Normality%20of%20residuals%20means%20that,errors%20are%20random%20and%20independent.)).

We hence determine normality using the `ols_plot_resid_hist()` of the olsrr package:

```{r}
ols_plot_resid_hist(price_mlr1)
```

::: {.callout-note style="color: steelblue"}
The figure above indicates that the residual fo the MLR model resembles a normal distribution.
:::

However, we also carry out further statistical test to ascertain this observation by using the `ols_test_normality()` function. We first extract the residuals, differences between the observed and predicted values, from the model using `residuals()`. We then apply the `sample()` function with the sample size set at 5000 as the `ols_test_normality()` function requires a sample size between 3 and 5000 and our data as-is would be too large as it comprises 11,642 observations.

```{r}
ols_test_normality(sample(residuals(price_mlr1), size = 5000))
```

::: {.callout-note style="color: steelblue"}
We note from the results that the p-values are smaller than alpha value of 0.05 at 95% confidence interval. We can hence reject the null hypothesis and determine that there is sufficient evidence that the residuals are not normally distributed.

Taking this result into consideration with the normality plot above, we can assess that the MLR model might not be able to fully capture the main patterns and sources of variation in the data.
:::

### 5.7.4 Testing for spatial autocorrelation

Spatial autocorrelation implies that observations that are near each other are more likely to be similar, which can violate the independence assumption in linear regression and lead to erroneous results. We will hence test for spatial autocorrelation.

We will first export the residual of the MLR model and save it as a data frame:

```{r}
price_mlr1.output <- as.data.frame(price_mlr1$residuals)
```

Next, we will join the newly created data frame with the *train_data* sf object:

```{r}
train_data_res <- cbind(train_data,
                        price_mlr1.output) %>%
  rename(`MLR_RES` = `price_mlr1.residuals`)
```

Next, we will use tmap to display the distribution of the residuals on an interactive map:

```{r}
#| code-fold: true
tmap_mode("view")
tm_shape(mpsz)+
  tmap_options(check.and.fix = TRUE)+
  tm_polygons(alpha = 0.4)+
  tm_shape(train_data_res)+
  tm_dots(col = "MLR_RES",
          alpha = 0.6,
          style = "quantile")+
  tm_view(set.zoom.limits = c(11,14))
```

We will switch the tmap mode back to "plot" using the code chunk below:

```{r}
tmap_mode("plot")
```

::: {.callout-note style="color: steelblue"}
The map above reveals that there seems to be signs of spatial autocorrelation as the model predicts higher than expected resale prices in the central and east regions of Singapore and lower than expected resale prices in the west and northeast regions.
:::

To determine if our observation is true, we will perform the Moran's I test. We will utilise the spdep package for the analysis.

First, we will convert the *train_data_res* sf data frame to a spatial point data frame using the `as_Spatial()` function and then compute the distance-based weight matrix using the `dnearneigh()` function of spdep package:

```{r}
#| eval: false
nb <- dnearneigh(coordinates(as_Spatial(train_data_res)), 0, 1500, longlat = FALSE)
```

We save the output as a new rds file to avoid re-running the step above:

```{r}
#| eval: false
write_rds(nb,"data/rds/nb.rds")
```

```{r}
nb <- read_rds("data/rds/nb.rds")
summary(nb)
```

Next, we will utilise `nb2listw()` of spdep package to convert the output neighbours list into spatial weights:

```{r}
nb_lw <- nb2listw(nb, style = 'W')
```

We then use the `lm.morantest()` function of spdep package to perform Moran's I test for residual spatial autocorrelation:

```{r}
#| eval: false
moran_mlr <- lm.morantest(price_mlr1, nb_lw)
```

We save the output as a new rds file to avoid re-running the step above and load it into the R environment:

```{r}
#| eval: false
write_rds(moran_mlr,"data/rds/moran_mlr.rds")
```

```{r}
moran_mlr <- read_rds("data/rds/moran_mlr.rds")
moran_mlr
```

::: {.callout-note style="color: steelblue"}
The Global Moran's I test for residual spatial autocorrelation shows that its p-value is less than 2.2e-16, which is less than the alpha value of 0.05 at 95% confidence interval. As such, we will reject the null hypothesis that residuals are randomly distributed. As the observed Global Moran I value is greater than 0, we can infer that the residuals resemble a cluster distribution.

Given this, we note that multiple linear regression might not be able to fully capture the varuiations in the data to predict HDB resale prices.
:::

### 5.8 Random Forest (RF)

In this section, we will build predictive models using RF. We will utilise functions from the [ranger](https://cran.r-project.org/web/packages/ranger/) (non-spatial RF) and [SpatialML](https://cran.r-project.org/web/packages/SpatialML/SpatialML.pdf) package.

### 5.8.1 Extracting coordinates data and dropping geometry field

Before we carry out further analysis, we will need to extract the x, y coordinates from the full *resale_tidy_all*, train and test data sets. We need to do this additional step for RF as functions under the SpatialML package are not able to use train and test data directly and take in the attribute data (in the form of a general data frame) and coordinates data as separate arguments.

The code chunk below extracts the x, y coordinates:

```{r}
coords_all <- st_coordinates(resale_tidy_all)
coords_train <- st_coordinates(train_data)
coords_test <- st_coordinates(test_data)
```

We save the output as new rds files:

```{r}
#| eval: false
write_rds(coords_all,"data/rds/coords_all.rds")
write_rds(coords_train,"data/rds/coords_train.rds")
write_rds(coords_test,"data/rds/coords_test.rds")
```

```{r}
coords_all <- read_rds("data/rds/coords_all.rds")
coords_train <- read_rds("data/rds/coords_train.rds")
coords_test <- read_rds("data/rds/coords_test.rds")
```

We drop the geometry column of the train and test sf data frames using `st_drop_geometry()` in the code chunk below:

```{r}
train_data_nogeom <- train_data %>%
  st_drop_geometry()

test_data_nogeom <- test_data %>%
  st_drop_geometry()
```

### 5.8.2 Non-spatial RF

We will utilise the `ranger()` function of the ranger package to create the rf model. To ensure reproducibility, we set a seed. We also indicate the number of trees to be 30, for a robust analysis while ensuring that it is not too computationally intensive, and will consistently use this across our rf analysis.

```{r}
#| eval: false
set.seed(1234)
rf <- ranger(resale_price ~ storey_range + floor_area_sqm + flat_model + remaining_lease_mth + PROX_ELDERCARE + PROX_HAWKER + PROX_PARK + PROX_SUPERMARKET + PROX_CLINIC + PROX_CHILDCARE + PROX_KINDERGARTEN + PROX_MRT + PROX_BUSSTOP + PROX_PRISCH + PROX_GOODPRISCH + PROX_CBD + PROX_MALL + WITHIN_350M_HAWKER + WITHIN_350M_SUPERMARKETS + WITHIN_350M_CHILDCARE + WITHIN_350M_KINDERGARTENS +  WITHIN_350M_BUSSTOP + WITHIN_1KM_PRISCH,
                data = train_data_nogeom,
                num.trees = 30)
```

We save the model as a new rds file and load it into the R environment to avoid re-running the step above:

```{r}
#| eval: false
write_rds(rf,"data/rds/rf.rds")
```

```{r}
rf <- read_rds("data/rds/rf.rds")
```

We view the report using the code chunk below:

```{r}
rf
```

::: {.callout-note style="color: steelblue"}
We note that the non-spatial RF model has an R2 value of 0.9113451 which means that the model is able to predict approximately 91.1% of HDB resale prices.

At present, based on the train data set, this is a better predictive performance than the conventional MLR model which was only able to predict 67.8% of 3-room HDB resale prices and comparable to the geographically weighted MLR model (fitted with train data) which was able to predict approximately 93.7% of HDB resale prices.
:::

### 5.8.3 Geographically Weighted RF (gwRF)

We will now calibrate the predictive model using the `grf()` function of SpatialML package. We will utilise an adaptive bandwidth as it would better be able to capture the local variations in the distribution of HDB resale units across Singapore.

#### 5.8.3.1 Computing adaptive bandwidth for train data

We will first determine the optimal bandwidth to run the grf model with by using the `grf.bw()` function of the SpatialML package below. We also keep to 30 trees for the analysis for consistency and to allow for comparison of results.

```{r}
#| eval: false
grf_bw_adaptive <- grf.bw(resale_price ~ storey_range + floor_area_sqm + flat_model + remaining_lease_mth + PROX_ELDERCARE + PROX_HAWKER + PROX_PARK + PROX_SUPERMARKET + PROX_CLINIC + PROX_CHILDCARE + PROX_KINDERGARTEN + PROX_MRT + PROX_BUSSTOP + PROX_PRISCH + PROX_GOODPRISCH + PROX_CBD + PROX_MALL + WITHIN_350M_HAWKER + WITHIN_350M_SUPERMARKETS + WITHIN_350M_CHILDCARE + WITHIN_350M_KINDERGARTENS +  WITHIN_350M_BUSSTOP + WITHIN_1KM_PRISCH,
                data = train_data_nogeom,
                kernel = "adaptive",
                coords = coords_train,
                bw.min = 20,
                bw.max = 100,
                step = 10,
                trees = 30)
```

We will save the results as a new rds to avoid re-running the above step as it can take a while:

```{r}
#| eval: false
write_rds(grf_bw_adaptive,"data/rds/grf_bw_adaptive.rds")
```

```{r}
grf_bw_adaptive <- read_rds("data/rds/grf_bw_adaptive.rds")
grf_bw_adaptive
```

Based on the result above, we note that the optimal bandwidth is 80.

#### 5.8.3.2 Constructing the adaptive bandwidth grf model

We then build the grf model using the adaptive bandwidth determined above. Likewise, we set the number of trees at 30. To ensure reproducibility, we set a seed.

```{r}
#| eval: false
set.seed(1234)
gwRF_adaptive_train <- grf(resale_price ~ storey_range + floor_area_sqm + flat_model + remaining_lease_mth + PROX_ELDERCARE + PROX_HAWKER + PROX_PARK + PROX_SUPERMARKET + PROX_CLINIC + PROX_CHILDCARE + PROX_KINDERGARTEN + PROX_MRT + PROX_BUSSTOP + PROX_PRISCH + PROX_GOODPRISCH + PROX_CBD + PROX_MALL + WITHIN_350M_HAWKER + WITHIN_350M_SUPERMARKETS + WITHIN_350M_CHILDCARE + WITHIN_350M_KINDERGARTENS +  WITHIN_350M_BUSSTOP + WITHIN_1KM_PRISCH,
                dframe = train_data_nogeom,
                bw = 80,
                kernel = "adaptive",
                coords = coords_train,
                ntree = 30)
```

We save the model output as a new rds file to avoid re-running the step above, since the process takes some time to run, and load it into the R environment:

```{r}
#| eval: false
write_rds(gwRF_adaptive_train,"data/rds/gwRF_adaptive_train.rds")
```

```{r}
gwRF_adaptive_train <- read_rds("data/rds/gwRF_adaptive_train.rds")
```

## 6. Comparison of model performance

To compare the performance of the three models, we will generate the predictions on the test data.

### 6.1 Predicting values of test data

### 6.1.1. Multiple Linear Regression (MLR)

We will utilise the *price_mlr1* model which has fitted with the *train_data* to make predictions on the test data using the `predict()` function in the code chunk below:

```{r}
test_data$predicted_mlr <- predict(price_mlr1,test_data)
```

### 6.1.2. Random Forest (RF)

For RF, we will also utilise the `predict()` function on the *rf* model which has fitted with the *train_data_nogeom* to make predictions on the test data in the code chunk below:

```{r}
set.seed(1234)
rf_pred <- predict(rf,
                   test_data_nogeom)
```

We then extract the predictions into a new field under the *test_data* data frame:

```{r}
test_data$predicted_rf <- rf_pred$predictions
```

### 6.1.3. Geographically Weighted RF (gwRF)

For gwRF, before we generate the predicted values of the test data, we first will need to prepare the test data set by combining it back with its coordinates data. We do so as the `predict.grf()` function requires the test data to be in the same data frame, different from the other functions under the SpatialML package used above.

The code chunk utilises `cbind()` to combine the test data with its corresponding coordinates data:

```{r}
test_data_gwrf <- cbind(test_data_nogeom,coords_test)
```

We then generate the predicted values of the test data using the code chunk below which utilises the `grf.predict()` function:

```{r}
#| eval: false
set.seed(1234)
gwRF_pred <- predict.grf(gwRF_adaptive_train,
                         test_data_gwrf,
                         x.var.name = "X",
                         y.var.name = "Y",
                         local.w = 1,
                         global.w = 0)
```

We save the output as a new rds file:

```{r}
#| eval: false
write_rds(gwRF_pred,"data/rds/gwRF_pred.rds")
```

```{r}
gwRF_pred <- read_rds("data/rds/gwRF_pred.rds")
```

The output of the `predict.grf()` is a vector of predicted values. We append the predicted values to the *test_data* sf data frame using the `cbind()` function in the code chunk below:

```{r}
test_data$predicted_gwrf <- gwRF_pred
```

We also save this overall updated *test_data* data frame as a new rds file:

```{r}
#| eval: false
write_rds(test_data,"data/rds/test_data_allres.rds")
```

```{r}
test_data_allres <- read_rds("data/rds/test_data_allres.rds")
```

### 6.2. Calculating R2 and Root Mean Square Error (RMSE)

To compare model performance, we will compute the folllowing metrics for each of the models:

-   [**Root Mean Square Error (RMSE)**]{.underline} which represents the standard deviation of the residuals (difference between predicted and actual resale prices): A lower RMSE indicates a model with more accurate predictions. This would allow us to determine which model can more accurately predict the HDB resale prices.

-   [**Residual analysis**]{.underline} complements the RMSE analysis by delving deeper into the accuracy of the model, specifically whether the models over- or under-predict.

-   [**Coefficient of determination (R2)**]{.underline} which indicates how well the independent variables in a statistical model explains the variation in the dependent variable: A higher R2 value suggests that the model and its chosen independent variables better capture the underlying variations in HDB resale prices. This would allow us to determine which combination of regression model and independent variables would be a better fit to predict HDB resale prices.

### 6.2.1 RMSE

#### 6.2.1.1 Overall RMSE values

We compute the RMSE values for each model using the `rmse()` function in the Metrics package. We then combine the results into a data frame using the `data.frame()` function.

```{r}
rmse_mlr <- Metrics::rmse(test_data_allres$resale_price,test_data_allres$predicted_mlr) 

rmse_rf <- Metrics::rmse(test_data_allres$resale_price,test_data_allres$predicted_rf)

rmse_gwrf <- Metrics::rmse(test_data_allres$resale_price,test_data_allres$predicted_gwrf) 

rmse_results <- data.frame(
  Model = c("Multiple Linear Regression","Random Forest", "Geographically Weighted Random Forest"),
  RMSE = c(rmse_mlr,rmse_rf,rmse_gwrf)
)

datatable(rmse_results)
```

::: {.callout-note style="color: steelblue"}
From the output above, we note that the geographically weighted random forest model has the lowest RMSE, followed by random forest and multiple linear regression. This indicates that the geographically weighted random model is better able to accurately predict the 3-room HDB resale prices as compared to the other two models.
:::

We can also plot a scatterplot to visualise the actual resale price and the predicted resale price using the code chunk below:

::: panel-tabset
## Multiple Linear Regression

```{r}
#| code-fold: true
ggplot(test_data_allres,
       aes(x=predicted_mlr,
           y=resale_price))+
  geom_point()+
  ggtitle("Scatterplot of Actual Resale Price vs Predicted Resale Price (MLR)")+
  labs(x="Predicted Resale Price",y="Actual Resale Price")
```

## Random Forest

```{r}
#| code-fold: true
ggplot(test_data_allres,
       aes(x=predicted_rf,
           y=resale_price))+
  geom_point()+
  ggtitle("Scatterplot of Actual Resale Price vs Predicted Resale Price (RF)")+
  labs(x="Predicted Resale Price",y="Actual Resale Price")
```

## Geographically Weighted Random Forest

```{r}
#| code-fold: true
ggplot(test_data_allres,
       aes(x=predicted_gwrf,
           y=resale_price))+
  geom_point()+
  ggtitle("Scatterplot of Actual Resale Price vs Predicted Resale Price (gwRF)")+
  labs(x="Predicted Resale Price",y="Actual Resale Price")
```
:::

::: {.callout-note style="color: steelblue"}
As with the result of the overall RMSE values determined above, we note that the geographically weighted random forest is the better predictive model as it has more scatter points closer to the diagonal line.
:::

#### 6.2.1.2 RMSE values by planning area

Beyond the overall RMSE values, we are also interested to find out the performance of the model across Singapore. We will do so at the planning area level by making use of the "PLN_AREA_N" field in the *mpsz* sf data object. We will first use `st_join()` to append the "PLN_AREA_N" attribute to the *test_data_allres* data frame in the code chunk below:

```{r}
test_data_area <- st_join(test_data_allres, mpsz["PLN_AREA_N"])
```

We then generate RMSE values based on "PLN_AREA_N" by first using `group_by()` to group the *test_data_area* data frame by "PLN_AREA_N", followed by determining the rmse values for each planning area and creating a new field "lowest rmse" for each planning area using the `case_when()` and `pmin()` function:

```{r}
#| eval: false
test_data_area <- test_data_area %>%
  group_by(PLN_AREA_N) %>%
  summarise(
    rmse_mlr = Metrics::rmse(resale_price,predicted_mlr),
    rmse_rf = Metrics::rmse(resale_price,predicted_rf),
    rmse_gwrf = Metrics::rmse(resale_price,predicted_gwrf)) %>%
  mutate(lowest_rmse = case_when(rmse_mlr == pmin(rmse_mlr,rmse_rf,rmse_gwrf) ~ "MLR",
                                 rmse_rf == pmin(rmse_mlr,rmse_rf,rmse_gwrf) ~ "RF"
,
rmse_gwrf == pmin(rmse_mlr,rmse_rf,rmse_gwrf) ~ "gwRF")) %>%
  st_drop_geometry()
```

We save the *test_data_area* data frame as a new rds file and load it into the environment for further analysis to avoid re-running the step above:

```{r}
#| eval: false
write_rds(test_data_area,"data/rds/test_data_area.rds")
```

```{r}
test_data_area <- read_rds("data/rds/test_data_area.rds")
```

In order to visualise best model with the lowest RMSE for each area, we will need to `left_join` the fields created in the above code chunk with the *mpsz* sf data object.

We will first utilise `group_by()` and `st_union()` to collapse the geometries within each planning area into a single, unified geometry and then utilise `left_join` to append the RMSE results created in the *test_data_area* data frame to the unified geometry created in *mpsz,* creating a new data frame *mpsz_test_rmse*.

```{r}
mpsz_test_rmse <- mpsz %>%
  group_by(PLN_AREA_N) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup() %>%
  left_join(test_data_area,by = "PLN_AREA_N")
```

We then utilise tmap to plot a high quality cartographic map displaying the best predictive model with the lowest RMSE for each planning area across Singapore:

```{r}
#| code-fold: true
tmap_mode("view")
tm_shape(mpsz_test_rmse)+
  tm_fill("lowest_rmse",
          title = "Lowest RMSE")+
  tm_borders(alpha = 0.5)+
  tm_compass(type = "8star",size = 2)+
  tm_grid(alpha = 0.2)+
  tm_layout(main.title = "Best predictive model with lowest RMSE for each planning area in Singapore",
            main.title.position = "center")+
  tm_credits("Source: Planning subzone boundary from Urban Redevelopment Authority",
             position = c("left","bottom"))+
  tm_view(set.zoom.limits = c(11,14))
```

::: {.callout-note style="color: steelblue"}
Based on the map above, we note that the geographically weighted random forest model is the best predictive model for most regions for the resale price of 3-room HDB resale flats, this is followed by the random forest model and lastly the multiple linear regression model.

Specifically,

-   the geographically weighted random forest model is better at predicting resale prices of 3-room HDB units in the central, east and part of the north (Woodlands) region of Singapore

-   the random forest model is better at predicting resale prices of 3-room HDB units in the west and peripheral regions of the central region

-   the multiple linear regression model is better at predicting resale prices of 3-room HDB units in part of the north (Sembawang) and north east region (Punggol and Sengkang)
:::

Before we proceed with further analysis, we switch the tmap_mode to "plot":

```{r}
tmap_mode("plot")
```

### 6.2.2 Residual Analysis

After we have determined the accuracy of the predictive models above, we will carry out residual analysis to better understand model performance in detail i.e. whether the model under- or over-predicts. We will carry out this analysis at the planning area level.

We will create a new sf data frame *test_data_residual* using the *test_data_allres* sf data frame created earlier and add new fields representing the residuals from each of the three models by using the `mutate()` function:

```{r}
#| eval: false
test_data_residual <- test_data_allres %>%
  mutate(residual_mlr = resale_price - predicted_mlr,
         residual_rf = resale_price - predicted_rf,
         residual_gwrf = resale_price - predicted_gwrf)
```

We will then use `st_join()` to append the “PLN_AREA_N” in the *mpsz* sf data object to the *test_data_residual* data frame in the code chunk below:

```{r}
#| eval: false
test_data_residual <- st_join(test_data_residual, mpsz["PLN_AREA_N"])
```

We will save this output as a new rds file to avoid re-running the above steps and load it into the R environment for further analysis:

```{r}
#| eval: false
write_rds(test_data_residual,"data/rds/test_data_residual.rds")
```

```{r}
test_data_residual <- read_rds("data/rds/test_data_residual.rds")
```

We then utilise tmap to plot a high quality cartographic map displaying the residuals for each model for each planning area across Singapore:

::: panel-tabset
## Multiple Linear Regression

```{r}
#| code-fold: true
tmap_mode("view")
tm_shape(mpsz)+
  tm_polygons(alpha = 0.1)+
  tm_shape(test_data_residual)+
  tm_dots(col = "residual_mlr",
          title = "Residual of MLR",
          breaks = seq(-400000, 600000, by = 100000),
          palette = "RdBu")+
  tm_compass(type = "8star",size = 2)+
  tm_grid(alpha = 0.2)+
  tm_layout(main.title = "Residual of MLR for each planning area in Singapore",
            main.title.position = "center")+
  tm_credits("Source: Planning subzone boundary from Urban Redevelopment Authority",
             position = c("left","bottom"))+
  tm_view(set.zoom.limits = c(11,14))
```

## Random Forest

```{r}
#| code-fold: true
tmap_mode("view")
tm_shape(mpsz)+
  tm_polygons(alpha = 0.1)+
  tm_shape(test_data_residual)+
  tm_dots(col = "residual_rf",
          title = "Residual of RF",
          breaks = seq(-400000, 600000, by = 100000),
          palette = "RdBu")+
  tm_compass(type = "8star",size = 2)+
  tm_grid(alpha = 0.2)+
  tm_layout(main.title = "Residual of RF for each planning area in Singapore",
            main.title.position = "center")+
  tm_credits("Source: Planning subzone boundary from Urban Redevelopment Authority",
             position = c("left","bottom"))+
  tm_view(set.zoom.limits = c(11,14))
```

## Geographically Weighted Random Forest

```{r}
#| code-fold: true
tmap_mode("view")
tm_shape(mpsz)+
  tm_polygons(alpha = 0.1)+
  tm_shape(test_data_residual)+
  tm_dots(col = "residual_gwrf",
          title = "Residual of gwRF",
          breaks = seq(-400000, 600000, by = 100000),
          palette = "RdBu")+
  tm_compass(type = "8star",size = 2)+
  tm_grid(alpha = 0.2)+
  tm_layout(main.title = "Residual of gwRF for each planning area in Singapore",
            main.title.position = "center")+
  tm_credits("Source: Planning subzone boundary from Urban Redevelopment Authority",
             position = c("left","bottom"))+
  tm_view(set.zoom.limits = c(11,14))
```
:::

::: {.callout-note style="color: steelblue"}
Based on the maps above, we note that:

-   the multiple linear regression model tends to under-predict in most regions of Singapore except for the west and northeast region where there are more over-predictions of resale price

-   the random forest and geographically weighted random forest models tend to under predict in most regions of Singapore, there seems to be less over-predictions as compared to the multiple linear regression model
:::

Before we proceed with further analysis, we switch the tmap_mode to "plot":

```{r}
tmap_mode("plot")
```

### 6.2.3 R2 values

#### 6.2.3.1 Overall R2 values

We compute the R2 values for each model using the `R2_Score()` function in the MLmetrics package. We then combine the results into a data frame using the `data.frame()` function.

```{r}
r2_mlr <- MLmetrics::R2_Score(test_data_allres$predicted_mlr,test_data_allres$resale_price) 

r2_rf <- MLmetrics::R2_Score(test_data_allres$predicted_rf,test_data_allres$resale_price) 

r2_gwrf <- MLmetrics::R2_Score(test_data_allres$predicted_gwrf,test_data_allres$resale_price)

r2_results <- data.frame(
  Model = c("Multiple Linear Regression","Random Forest", "Geographically Weighted Random Forest"),
  R2_Score = c(r2_mlr,r2_rf,r2_gwrf)
)

datatable(r2_results)
```

::: {.callout-note style="color: steelblue"}
From the output above, we note that the geographically weighted random forest model has the highest R2 score, followed by random forest and multiple linear regression.

This indicates that the geographically weighted random forest model and its independent variables are better able to capture the underlying variations in HDB resale prices.

To recap, all models used the same independent variables with the exception of multiple linear regression as PROX_BUSSTOP, PROX_PRISCH, PROX_MALL and WITHIN_350M_BUSSTOP were not statistically significant for the multiple linear regression model and removed.
:::

#### 6.2.3.2 R2 values by planning area

Beyond the overall R2 values, we are also interested to find out how the R2 score of the models differ across Singapore. We will do so at the planning area level by making use of the "PLN_AREA_N" field in the *mpsz* sf data object. We will first use `st_join()` to append the "PLN_AREA_N" attribute to the *test_data_allres* data frame in the code chunk below:

```{r}
test_data_area_r2 <- st_join(test_data_allres, mpsz["PLN_AREA_N"])
```

We then generate R2 scores based on "PLN_AREA_N" by first using `group_by()` to group the *test_data_area_r2* data frame by "PLN_AREA_N", followed by determining the R2 score for each planning area and creating a new field "best r2" for each planning area using the `case_when()` and `pmax()` function:

```{r}
#| eval: false
test_data_area_r2 <- test_data_area_r2 %>%
  group_by(PLN_AREA_N) %>%
  summarise(
    r2_mlr = MLmetrics::R2_Score(predicted_mlr,resale_price),
    r2_rf = MLmetrics::R2_Score(predicted_rf,resale_price),
    r2_gwrf = MLmetrics::R2_Score(predicted_gwrf,resale_price)) %>%
  mutate(highest_r2 = case_when(r2_mlr == pmax(r2_mlr,r2_rf,r2_gwrf) ~ "MLR",
                                 r2_rf == pmax(r2_mlr,r2_rf,r2_gwrf) ~ "RF"
,
r2_gwrf == pmax(r2_mlr,r2_rf,r2_gwrf) ~ "gwRF")) %>%
  st_drop_geometry()
```

We save the *test_data_area_r2* data frame as a new rds file and load it into the environment for further analysis to avoid re-running the step above:

```{r}
#| eval: false
write_rds(test_data_area_r2,"data/rds/test_data_area_r2.rds")
```

```{r}
test_data_area_r2 <- read_rds("data/rds/test_data_area_r2.rds")
```

In order to visualise best model with the highest R2 for each area, we will need to `left_join` the fields created in the above code chunk with the *mpsz* sf data object.

We will first utilise `group_by()` and `st_union()` to collapse the geometries within each planning area into a single, unified geometry and then utilise `left_join` to append the R2 results created in the *test_data_area_r2* data frame to the unified geometry created in *mpsz,* creating a new data frame *mpsz_test_r2*.

```{r}
mpsz_test_r2 <- mpsz %>%
  group_by(PLN_AREA_N) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup() %>%
  left_join(test_data_area_r2,by = "PLN_AREA_N")
```

We then utilise tmap to plot a high quality cartographic map displaying the predictive model with the highest R2 score for each planning area across Singapore:

```{r}
#| code-fold: true
tmap_mode("view")
tm_shape(mpsz_test_r2)+
  tm_fill("highest_r2",
          title = "Highest R2")+
  tm_borders(alpha = 0.5)+
  tm_compass(type = "8star",size = 2)+
  tm_grid(alpha = 0.2)+
  tm_layout(main.title = "Predictive model with highest R2 for each planning area in Singapore",
            main.title.position = "center")+
  tm_credits("Source: Planning subzone boundary from Urban Redevelopment Authority",
             position = c("left","bottom"))+
  tm_view(set.zoom.limits = c(11,14))
```

::: {.callout-note style="color: steelblue"}
Based on the map above, we note that the R2 scores/goodness-of-fit of the geographically weighted random forest model for the resale price of 3-room HDB resale flats is the best for most regions, this is followed by the random forest model and lastly the multiple linear regression model.

Specifically,

-   the goodness-of-fit of the geographically weighted random forest model is better in the central, part of the east (Tampines) and part of the north (Woodlands) region of Singapore

-   the goodness-of-fit of the random forest model is better in the west and peripheral regions of the central region

-   the goodness-of-fit of the multiple linear regression model is better in part of the north (Sembawang), northeast (Punggol and Sengkang) and east region (Pasir Ris and Changi)
:::

Note that we will need switch the tmap_mode to "plot" to save bandwidth:

```{r}
tmap_mode("plot")
```

#### 6.2.3.3 Variable importance for geographically weighted random forest

The geographically weighted random forest model is the most promising predictive model among the three models studied in this exercise as it has the overall lowest RMSE and highest R2 score in predicting the resale prices of 3-room HDB flats across Singapore.

We will hence further determine the order of importance of the independent variables of the geographically weighted random forest by utilising the output feature class table *gwRF_adaptive_train* generated in section [5.8.3.2 Constructing the adaptive bandwidth grf model]. We do so by:

-   extracting the Global Model "variable.importance" into a data frame

-   creating a new field "Variable" using `rownames_to_column()` to house the names of the independent variables

-   renaming the variable importance variable to a shorter and more user-friendly term "Value"

-   arrange the variables in descending order of variable importance

-   use `scales()` to format the values (by adding commas) to be more readable as the numbers are very long

```{r}
gwRF_varorder <- as.data.frame(gwRF_adaptive_train$Global.Model$variable.importance) %>%
  rownames_to_column(var = "Variable") %>%
  rename("Value" =  "gwRF_adaptive_train$Global.Model$variable.importance") %>%
  arrange(desc(Value)) %>%
  mutate(Value = scales::comma(Value))
                               
datatable(gwRF_varorder)
```

::: {.callout-note style="color: steelblue"}
From the output above, we note that structural factors are more important than locational factors, with the remaining lease as the most important factor. Notably, the value of variable importance for "remaining_lease_mth" is more than twice the value of the next most important variable.

Among the locational factors, we note that proximity to CBD is the most important, followed by proximity to MRT, hawker centres, parks, shopping malls, eldercare services, kindergartens, good primary schools, primary schools, and CHAS clinics. This is followed by the number of hawker centres within 350m (aka walking distance) and number of primary schools within 1 km of the HDB units. The remaining variables have variable importance of less than 1 million and will not be elaborated in detail.

The order of variable importance could potentially be explained by the demographics of the population who are interested in 3-room HDB flats, as they may be smaller unit families that have no to few children, thus contributing to the lower ranking of locational factors related to schools.
:::

## 7. Conclusion

Overall, we observe that the geographically weighted random forest model to be a better predictive model as compared to the two other models studied in this exercise, multiple linear regression and random forest as it has the lowest overall RMSE and has the lowest RMSE for most regions in Singapore for the resale prices of 3-room HDB flats. Further, it has the highest overall R2 score which indicates that the model and its independent variables are better able to capture the underlying variations in HDB resale prices as compared to the other two models. This indicates the importance of taking in the spatial autocorrelation and spatial heterogeneity in geographic datasets such as HDB resale transactions.

We also observe structural factors to have greater variable importance as compared to locational factors. For future work, we could study more structural factors and how they might influence the predictive power of the model. Potential factors could include the last renovation date of the HDB flat as a recently renovated flat might attract higher prices, given that less renovation might need to be done for the new buyer to move in, and also whether the flat is west-facing as buyers could be concerned about high temperatures/hot afternoon sun.

## 8. References

-   <https://r4gdsa.netlify.app/>

-   <https://pro.arcgis.com/en/pro-app/latest/tool-reference/spatial-statistics/how-geographicallyweightedregression-works.htm>

-   <https://pmc.ncbi.nlm.nih.gov/articles/PMC11205154/>

-   <https://www.asiaone.com/money/more-3-room-resale-flats-change-hands-800k-and-higher-2024>

-   <https://www.moe.gov.sg/primary/p1-registration/distance>

-   <https://www.linkedin.com/advice/0/what-best-methods-assess-normality-residuals#:~:text=Normality%20of%20residuals%20means%20that,errors%20are%20random%20and%20independent.>
