---
title: "In-class Exercise 9"
author: "Kock Si Min"
date: October 28, 2024
date-modified: "last-modified"
toc: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

# Modelling Geographical Accessibility

-   Prof shared that for real-life cases, we cannot use fixed capacity and demand i.e. 100 used in hands-on exercise 9, instead we need to determine real figures

## The Data

Four datasets will be used:

-   `MP14_SUBZONE_NO_SEA_PL`: URA Master Plan 2014 subzone boundary GIS data. This data set is downloaded from data.gov.sg.

-   `hexagons`: A 250m radius hexagons GIS data. This data set was created by using [*st_make_grid()*](https://r-spatial.github.io/sf/reference/st_make_grid.html) of sf package. It is in ESRI shapefile format.

-   `ELDERCARE`: GIS data showing location of eldercare service. [This data](https://data.gov.sg/dataset/eldercare-services) is downloaded from data.gov.sg. There are two versions. One in ESRI shapefile format. The other one in Google kml file format. For the purpose of this hands-on exercise, ESRI shapefile format is provided.

-   `OD_Matrix`: a distance matrix in csv format. There are six fields in the data file. They are:

    -   `origin_id`: the unique id values of the origin (i.e. `fid` of hexagon data set.),

    -   `destination_id`: the unique id values of the destination (i.e. `fid` of `ELDERCARE` data set.),

    -   `entry_cost`: the perpendicular distance between the origins and the nearest road),

    -   `network_cost`: the actual network distance from the origin and destination,

    -   `exit_cost`: the perpendicular distance between the destination and the nearest road), and

    -   `total_cost`: the summation of `entry_cost`, `network_cost` and `exit_cost`.

All values of the cost-related fields are in metres.

## Getting Started

```{r}
pacman::p_load(sf,SpatialAcc,tidyverse,tmap,ggstatsplot, reshape2,rstantools)
```

## Geospatial Data Wrangling

### Importing geospatial data

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MP14_SUBZONE_NO_SEA_PL") %>%
  st_transform(3414)
```

```{r}
hexagons <- st_read(dsn = "data/geospatial",
                layer = "hexagons") %>%
  st_transform(3414)
```

```{r}
eldercare <- st_read(dsn = "data/geospatial",
                layer = "ELDERCARE") %>%
  st_transform(3414)
```

### Data cleaning and updating attribute fields of the geospatial data

-   supply

    -   artificially create a new column "capacity" to make it easy to calibrate the model

    -   but in reality, need to collect real data i.e. survey, web crawling to collect actual capacity of eldercare centre

    -   retain fid and ADDRESSPOS which represent id and address of eldercare

```{r}
eldercare <- eldercare %>%
  select(fid,ADDRESSPOS) %>%
  mutate(capacity = 100)
```

-   demand

    -   artificially create a new column "demand" to make it easy to calibrate the model

    -   but in reality, need to collect real data i.e. crawl from hdb website the number of floors then multiply by constant unit to estimate number of elderly folks

    -   retain unique id aka fid

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

## Aspatial Data Handling and Wrangling

### Importing Distance Matrix

-   for info, Prof used [r5r](https://cran.r-project.org/web/packages/r5r/vignettes/r5r.html) to create ODMatrix; is a java library that calculate shorter distance from road network i.e. OpenStreetMap

-   note that ODMatrix is in the structure of long and thin - this is how its returned from r5r:

    -   total cost is summation of entry, network and exit costs

    -   entry cost: perpendicular distance from origin point to nearest road i.e. walk/bus to mrt distance

    -   exit cost: perpendicular distance from destination point to nearest road

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

### Tidying distance matrix

-   **rows represent origins** i.e. aka **the from field** and the **columns represent destinations** i.e. aka the to field
-   code chunk below uses *spread()* of **tidyr** package to **transform the ODMatrix from a thin format into a fat format:**

```{r}
distmat <- ODMatrix %>%
  select(origin_id,destination_id,total_cost) %>%
  pivot_wider(names_from= destination_id,
              values_from = total_cost) %>%
  select(c(-c('origin_id')))
```

```{r}
distmat_km <- as.matrix(distmat/1000)
```

-   note that distmat_km is 3125 observations of 120 variables - 3125 match with hexagons 3125 observations of 3 variables (demand), 120 match with eldercare 120 observations of 4 variables (supply)

-   Prof shared that should use pivot_longer()

## Modelling and Visualising Accessibility using Hansen Method

### Computing Hansen's accessibility

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km,
                            d0 = 50,
                            power = 2,
                            family = "Hansen"))

colnames(acc_Hansen) <- "accHansen"
acc_Hansen <- as_tibble(acc_Hansen)
hexagon_Hansen <- bind_cols(hexagons,acc_Hansen)
```

### Visualising Hansen's accessibility

#### Extracting map extend

We will extract the extend of hexagons simple feature data frame by using *st_bbox()* of sf package:

```{r}
mapex <- st_bbox(hexagons)
mapex
```

The code chunk below uses a collection of mapping functions of the **tmap** package to create a high cartographic quality map to indicate the accessibility to eldercare centres in Singapore:

```{r}
tmap_mode('plot')
tm_shape(hexagon_Hansen,
         bbox = mapex) +
  tm_fill(col = "accHansen",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1)+
  tm_shape(eldercare) +
  tm_symbols(size = 0.1, col = "skyblue") +
  tm_layout(main.title = "Accessibility to eldercare: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.28,
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right","top"),
            frame = TRUE) +
  tm_compass(type = "8star", size = 2) +
  tm_scale_bar(width = 0.15)+
  tm_grid(lwd = 0.1, alpha = 0.5)
```

::: callout-note
## Questions to raise

-   why do i need to indicate bbox = mapex when the geometry in hexagon_Hansen is based on the hexagons polygon geometry?
    -   Prof wants to make sure extend does not change, ensure that every plot will have the same map extend
:::

### Statistical graphic visualisation

```{r}
hexagon_Hansen <- st_join(hexagon_Hansen, mpsz,
                          join = st_intersects)
```

We use ggplot to plot the distribution via boxplot:

```{r}
ggplot(hexagon_Hansen,
       aes(y=log(accHansen),
           x= REGION_N))+
  geom_boxplot()+
  geom_point(stat = "summary",
             fun.y = "mean",
             color = "red",
             size = 2)
```

We also utilise ggbetweenstats:

-   combines EDA with CDA and showcases it in one framework

-   type = "p" indicate parametric test which compares mean value, "np" would mean non-parametric test which compares the median, "robust" method would trim outliers, "bayes" method means it create multiple permutations before presenting cut-off line

```{r}
ggbetweenstats(data = hexagon_Hansen,
               x= REGION_N,
               y = accHansen,
               type = "p")
```

## 
